---
title: "Long Island Accident Data Cleaning Walkthrough"
author: "Vraj Patel"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(data.table)
library(knitr)

mode_value <- function(x) {
  observed <- x[!is.na(x)]
  if (!length(observed)) {
    return(NA)
  }

  counts <- sort(table(observed), decreasing = TRUE)
  names(counts)[1]
}

median_datetime <- function(x, tz = "UTC") {
  observed <- x[!is.na(x)]
  if (!length(observed)) {
    return(as.POSIXct(NA))
  }

  as.POSIXct(median(as.numeric(observed)), origin = "1970-01-01", tz = tz)
}

missing_rate <- function(dt) {
  data.table(
    column = names(dt),
    missing_rate = vapply(dt, function(x) mean(is.na(x)), numeric(1))
  )[order(-missing_rate, column)]
}

row_audit <- data.table(
  step = character(),
  rows_before = integer(),
  rows_after = integer(),
  rows_removed = integer(),
  notes = character()
)

data_path <- "US_Accidents_March23.csv"
clean_path <- "li_accidents_clean.csv"
audit_path <- "li_cleaning_audit.csv"
decision_path <- "li_cleaning_decisions.csv"
```

This report keeps the raw national file unchanged, applies the Long Island ZIP filter, and then cleans the subset in clear, reproducible steps. The final shared dataset is written to `li_accidents_clean.csv`.

# Step 1: Raw data and Long Island ZIP filter

```{r read-raw-data}
raw_data <- fread(data_path, showProgress = FALSE)

original_rows <- nrow(raw_data)
original_cols <- ncol(raw_data)

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Raw national file",
    rows_before = original_rows,
    rows_after = original_rows,
    rows_removed = 0L,
    notes = "Original March 2023 US accidents file."
  )
)

raw_overview <- data.table(
  metric = c("Rows", "Columns"),
  value = c(original_rows, original_cols)
)

kable(raw_overview, caption = "Raw dataset overview")
```

```{r filter-long-island-zips}
raw_data[, nzip := suppressWarnings(as.numeric(Zipcode))]

li_data <- raw_data[
  (nzip >= 11000 & nzip < 11100) |
    (nzip >= 11500 & nzip < 11600) |
    (nzip >= 11700 & nzip < 12000)
]

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Long Island ZIP filter",
    rows_before = original_rows,
    rows_after = nrow(li_data),
    rows_removed = original_rows - nrow(li_data),
    notes = "Kept ZIP codes in the requested 110xx, 115xx, and 117xx-119xx ranges."
  )
)

filter_counts <- row_audit[step %in% c("Raw national file", "Long Island ZIP filter"),
                           .(step, rows_after, rows_removed, notes)]
kable(filter_counts, caption = "Row counts before and after the geographic filter")
```

```{r validate-geographic-filter}
zip_spotcheck <- li_data[!is.na(nzip), .N, by = .(Zipcode, nzip)][order(nzip)]
zip_spotcheck <- zip_spotcheck[c(1, 2, 3, .N - 2, .N - 1, .N)]

county_validation <- li_data[, .N, by = County][order(-N)]
outside_primary_counties <- li_data[
  !(County %in% c("Nassau", "Suffolk")),
  .N,
  by = .(County, Zipcode)
][order(County, Zipcode)]

if (!nrow(outside_primary_counties)) {
  outside_primary_counties <- data.table(
    County = "None",
    Zipcode = "None",
    N = 0L
  )
}

kable(zip_spotcheck, caption = "ZIP spot-check from the filtered subset")
kable(county_validation, caption = "County distribution after the ZIP filter")
kable(outside_primary_counties, caption = "Rows outside Nassau/Suffolk after ZIP filtering")
```

The requested ZIP filter works and produces a Long Island-focused subset. The check after filtering shows that most of the records are from Nassau and Suffolk, with a small number from Queens because some border ZIP codes fall in the same range. That result is documented instead of being changed silently.

\newpage

# Step 2: Conservative cleaning pipeline

```{r remove-exact-duplicates}
rm(raw_data)
invisible(gc())

clean_data <- copy(li_data)
rm(li_data)
invisible(gc())

before_duplicates <- nrow(clean_data)
clean_data <- unique(clean_data)
after_duplicates <- nrow(clean_data)

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Remove exact duplicates",
    rows_before = before_duplicates,
    rows_after = after_duplicates,
    rows_removed = before_duplicates - after_duplicates,
    notes = "Removed only full-row duplicates."
  )
)

kable(row_audit[step == "Remove exact duplicates", .(step, rows_before, rows_after, rows_removed)],
      caption = "Exact duplicate removal")
```

```{r drop-high-missing-columns}
before_high_missing <- nrow(clean_data)
initial_missing_profile <- missing_rate(clean_data)
high_missing_cols <- initial_missing_profile[missing_rate > 0.70, column]

if (length(high_missing_cols)) {
  clean_data <- clean_data[, !..high_missing_cols]
}

removed_high_missing <- if (length(high_missing_cols)) {
  data.table(column = high_missing_cols, action = "Dropped (>70% missing)")
} else {
  data.table(column = "None", action = "No columns exceeded the 70% missingness threshold")
}

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Drop >70% missing columns",
    rows_before = before_high_missing,
    rows_after = nrow(clean_data),
    rows_removed = 0L,
    notes = paste("Columns removed:", if (length(high_missing_cols)) paste(high_missing_cols, collapse = ", ") else "none")
  )
)

kable(removed_high_missing, caption = "High-missing columns assessment")
```

```{r handle-missing-values}
critical_columns <- intersect(
  c("ID", "Severity", "Start_Time", "End_Time", "Start_Lat", "Start_Lng",
    "Street", "City", "County", "State", "Zipcode", "Timezone"),
  names(clean_data)
)

supplementary_columns <- setdiff(names(clean_data), critical_columns)
missing_profile <- missing_rate(clean_data)

imputation_log <- data.table(
  column = character(),
  missing_before = integer(),
  fill_method = character(),
  fill_value = character()
)

for (column_name in critical_columns) {
  missing_before <- sum(is.na(clean_data[[column_name]]))
  if (!missing_before) {
    next
  }

  column_data <- clean_data[[column_name]]

  if (inherits(column_data, "POSIXct")) {
    fill_value <- median_datetime(column_data)
    fill_method <- "Median datetime"
  } else if (is.numeric(column_data) || is.integer(column_data)) {
    fill_value <- median(column_data, na.rm = TRUE)
    fill_method <- "Median"
  } else {
    fill_value <- mode_value(column_data)
    fill_method <- "Mode"
  }

  clean_data[is.na(get(column_name)), (column_name) := fill_value]

  imputation_log <- rbind(
    imputation_log,
    data.table(
      column = column_name,
      missing_before = missing_before,
      fill_method = fill_method,
      fill_value = as.character(fill_value)
    )
  )
}

if (!nrow(imputation_log)) {
  imputation_log <- data.table(
    column = "None",
    missing_before = 0L,
    fill_method = "No critical imputations needed",
    fill_value = "Critical fields were complete after filtering"
  )
}

missing_profile <- missing_rate(clean_data)
low_missing_supp_cols <- missing_profile[
  column %in% supplementary_columns & missing_rate > 0 & missing_rate < 0.05,
  column
]

before_low_missing_drop <- nrow(clean_data)
if (length(low_missing_supp_cols)) {
  clean_data <- clean_data[complete.cases(clean_data[, ..low_missing_supp_cols])]
}
after_low_missing_drop <- nrow(clean_data)

flagged_columns <- missing_rate(clean_data)[
  column %in% supplementary_columns & missing_rate >= 0.05,
  .(column, missing_rate, action = "Retained and documented")
]

if (!nrow(flagged_columns)) {
  flagged_columns <- data.table(
    column = "None",
    missing_rate = 0,
    action = "No supplementary columns required flagging"
  )
}

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Handle missing values",
    rows_before = before_low_missing_drop,
    rows_after = after_low_missing_drop,
    rows_removed = before_low_missing_drop - after_low_missing_drop,
    notes = paste(
      "Critical fields imputed when needed;",
      if (length(low_missing_supp_cols)) {
        paste("dropped rows missing:", paste(low_missing_supp_cols, collapse = ", "))
      } else {
        "no low-missing supplementary row drops were needed"
      }
    )
  )
)

kable(imputation_log, caption = "Critical-field imputation log")
kable(
  data.table(column = if (length(low_missing_supp_cols)) low_missing_supp_cols else "None"),
  caption = "Supplementary columns that triggered row drops (<5% missing)"
)
kable(flagged_columns, caption = "Supplementary columns retained with documented higher missingness")
```

```{r standardize-types}
before_type_standardization <- nrow(clean_data)

clean_data[, nzip := as.integer(nzip)]
clean_data[, Severity := factor(as.integer(Severity), levels = 1:4, ordered = TRUE)]

time_columns <- intersect(c("Start_Time", "End_Time", "Weather_Timestamp"), names(clean_data))
for (column_name in time_columns) {
  parsed_time <- as.POSIXct(clean_data[[column_name]], format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
  set(clean_data, j = column_name, value = parsed_time)
}

clean_data[, Duration_Minutes := as.numeric(difftime(End_Time, Start_Time, units = "mins"))]

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Standardize data types",
    rows_before = before_type_standardization,
    rows_after = nrow(clean_data),
    rows_removed = 0L,
    notes = "Parsed date/time fields, kept numeric ZIP in nzip, and converted Severity to an ordered factor."
  )
)

type_summary <- data.table(
  field = c("nzip", "Severity", "Start_Time", "End_Time", "Weather_Timestamp", "Duration_Minutes"),
  type_after_cleaning = c(
    class(clean_data$nzip)[1],
    class(clean_data$Severity)[1],
    class(clean_data$Start_Time)[1],
    class(clean_data$End_Time)[1],
    class(clean_data$Weather_Timestamp)[1],
    class(clean_data$Duration_Minutes)[1]
  )
)

kable(type_summary, caption = "Key standardized variable types")
```

```{r remove-implausible-values}
before_implausible_filter <- nrow(clean_data)

implausible_checks <- data.table(
  check = c(
    "Severity outside 1-4",
    "Negative duration",
    "Negative distance",
    "Start latitude outside [-90, 90]",
    "Start longitude outside [-180, 180]",
    "End latitude outside [-90, 90]",
    "End longitude outside [-180, 180]"
  ),
  failures = c(
    sum(!(as.integer(clean_data$Severity) %in% 1:4)),
    sum(!is.na(clean_data$Duration_Minutes) & clean_data$Duration_Minutes < 0),
    sum(!is.na(clean_data[["Distance(mi)"]]) & clean_data[["Distance(mi)"]] < 0),
    sum(clean_data$Start_Lat < -90 | clean_data$Start_Lat > 90, na.rm = TRUE),
    sum(clean_data$Start_Lng < -180 | clean_data$Start_Lng > 180, na.rm = TRUE),
    sum(!is.na(clean_data$End_Lat) & (clean_data$End_Lat < -90 | clean_data$End_Lat > 90)),
    sum(!is.na(clean_data$End_Lng) & (clean_data$End_Lng < -180 | clean_data$End_Lng > 180))
  )
)

clean_data <- clean_data[
  as.integer(Severity) %in% 1:4 &
    (is.na(Duration_Minutes) | Duration_Minutes >= 0) &
    (is.na(`Distance(mi)`) | `Distance(mi)` >= 0) &
    between(Start_Lat, -90, 90) &
    between(Start_Lng, -180, 180) &
    (is.na(End_Lat) | between(End_Lat, -90, 90)) &
    (is.na(End_Lng) | between(End_Lng, -180, 180))
]

after_implausible_filter <- nrow(clean_data)

row_audit <- rbind(
  row_audit,
  data.table(
    step = "Remove implausible values",
    rows_before = before_implausible_filter,
    rows_after = after_implausible_filter,
    rows_removed = before_implausible_filter - after_implausible_filter,
    notes = "Removed records only when values were logically impossible."
  )
)

kable(implausible_checks, caption = "Implausible-value checks")
```

\newpage

# Step 3: Outputs and documentation

```{r export-clean-data}
export_data <- copy(clean_data)
export_data[, Severity := as.integer(as.character(Severity))]

for (column_name in time_columns) {
  formatted_time <- format(export_data[[column_name]], "%Y-%m-%d %H:%M:%S", tz = "UTC")
  formatted_time[is.na(export_data[[column_name]])] <- NA_character_
  set(export_data, j = column_name, value = formatted_time)
}

fwrite(export_data, clean_path)

final_summary <- data.table(
  metric = c("Final cleaned rows", "Final cleaned columns", "Output file"),
  value = c(nrow(export_data), ncol(export_data), clean_path)
)

kable(final_summary, caption = "Cleaned dataset written for team use")
```

```{r cleaning-summary}
summary_table <- copy(row_audit)
summary_table[, final_dataset_rows := nrow(export_data)]

decision_log <- data.table(
  decision = c(
    "Geographic scope",
    "Duplicate handling",
    "High-missing columns",
    "Critical missingness",
    "Low-missing supplementary fields",
    "Higher-missing supplementary fields",
    "Implausible values"
  ),
  rationale = c(
    "Applied the requested ZIP ranges and checked the results with ZIP and county counts.",
    "Removed only exact duplicate rows so different accident records were not merged by mistake.",
    "Used a >70% missingness threshold, and no columns crossed it in the Long Island subset.",
    "Set up the report to fill in critical fields if needed, but none were missing after filtering.",
    "Dropped rows only when a supplementary field had very little missing data and removing those rows was reasonable.",
    "Kept End_Lat, End_Lng, Precipitation(in), and Wind_Chill(F) because they may still be useful even with missing values.",
    "Removed rows only when values were clearly impossible, such as negative duration or invalid coordinates."
  )
)

fwrite(summary_table, audit_path)
fwrite(decision_log, decision_path)

kable(summary_table, caption = "Row-count audit trail")
kable(decision_log, caption = "Plain-language rationale for each cleaning decision")
```

The final cleaned file is ready for shared use across the research questions. The raw CSV stays unchanged, the workflow stays reproducible from this single R Markdown file, and the validation shows clearly what was cleaned and what was kept.
