---
title: "Long Island Clustering Workflow"
author: "AMS 597 Project Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r helpers, include=FALSE}
ordered_weekdays <- c(
  "Monday", "Tuesday", "Wednesday", "Thursday",
  "Friday", "Saturday", "Sunday"
)

weather_group_map <- list(
  Clear = c("Fair", "Clear", "Fair / Windy"),
  Cloudy = c(
    "Mostly Cloudy", "Partly Cloudy", "Scattered Clouds", "Overcast", "Cloudy",
    "Mostly Cloudy / Windy", "Partly Cloudy / Windy", "Cloudy / Windy"
  ),
  Rain = c(
    "Light Rain", "Rain", "Heavy Rain", "Light Drizzle", "Light Rain / Windy",
    "Heavy Rain / Windy", "Drizzle"
  ),
  Snow_Ice = c(
    "Light Snow", "Snow", "Heavy Snow", "Wintry Mix",
    "Light Freezing Rain", "Ice Pellets", "Sleet"
  ),
  Fog_Haze = c("Fog", "Haze", "Smoke", "Mist", "Patches of Fog", "Shallow Fog")
)

normalize_wind_direction <- function(x) {
  x <- trimws(as.character(x))
  x_upper <- toupper(x)

  mapped <- dplyr::case_when(
    x_upper %in% c("N", "NORTH") ~ "N",
    x_upper %in% c("NE", "NORTHEAST", "NORTH EAST") ~ "NE",
    x_upper %in% c("E", "EAST") ~ "E",
    x_upper %in% c("SE", "SOUTHEAST", "SOUTH EAST") ~ "SE",
    x_upper %in% c("S", "SOUTH") ~ "S",
    x_upper %in% c("SW", "SOUTHWEST", "SOUTH WEST") ~ "SW",
    x_upper %in% c("W", "WEST") ~ "W",
    x_upper %in% c("NW", "NORTHWEST", "NORTH WEST") ~ "NW",
    x_upper %in% c("CALM", "VARIABLE", "VAR") ~ "Variable_Calm",
    TRUE ~ "Variable_Calm"
  )

  factor(
    mapped,
    levels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "Variable_Calm")
  )
}

mode_value <- function(x) {
  values <- x[!is.na(x)]
  if (!length(values)) {
    return(NA_character_)
  }

  names(sort(table(values), decreasing = TRUE))[1]
}
```

# Phase 1: Environment and Reproducibility Setup

This file is the main reproducible workflow for the Long Island clustering analysis. It is meant to run from a clean R session and use `li_accidents_clean.csv` as the only project dataset input.

```{r phase-1-config}
required_packages <- c(
  "data.table",
  "dplyr",
  "lubridate",
  "clustMixType",
  "dbscan",
  "cluster",
  "fpc",
  "ggplot2",
  "geosphere",
  "factoextra",
  "knitr",
  "rmarkdown"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste(
      "Install these packages before running the full workflow:",
      paste(missing_packages, collapse = ", ")
    )
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

set.seed(42)

input_path <- "../li_accidents_clean.csv"
output_dir <- "local_artifacts/output"

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

if (!file.exists(input_path)) {
  stop("Expected input file not found: ", input_path)
}

phase_1_check <- data.frame(
  item = c(
    "Input dataset",
    "Output directory",
    "Seed",
    "Working directory"
  ),
  value = c(
    normalizePath(input_path, winslash = "/"),
    normalizePath(output_dir, winslash = "/"),
    "42",
    normalizePath(".", winslash = "/")
  )
)

knitr::kable(phase_1_check, caption = "Phase 1 setup check")
```

```{r load-data-preview}
df_raw <- data.table::fread(input_path, showProgress = FALSE)

data_overview <- data.frame(
  rows = nrow(df_raw),
  columns = ncol(df_raw),
  source_file = basename(input_path)
)

knitr::kable(data_overview, caption = "Input file loaded successfully")
```

# Phase 2: Data Ingestion and Raw Profiling

This phase stays read-only. The goal is to confirm that the cleaned Long Island dataset matches the expectations in the execution plan before any feature engineering starts.

```{r phase-2-dimension-checks}
expected_rows <- 33375L
expected_cols <- 48L

if (nrow(df_raw) != expected_rows || ncol(df_raw) != expected_cols) {
  stop(
    paste0(
      "Unexpected dataset dimensions. Expected ",
      expected_rows, " rows and ", expected_cols, " columns, got ",
      nrow(df_raw), " rows and ", ncol(df_raw), " columns."
    )
  )
}

phase_2_dimensions <- data.frame(
  metric = c("Rows", "Columns"),
  expected = c(expected_rows, expected_cols),
  observed = c(nrow(df_raw), ncol(df_raw))
)

knitr::kable(phase_2_dimensions, caption = "Phase 2 dimension check")
```

```{r phase-2-county-and-severity}
county_profile <- as.data.table(df_raw)[, .N, by = County][order(match(County, c("Nassau", "Suffolk", "Queens")))]
severity_profile <- as.data.table(df_raw)[, .N, by = Severity][order(as.integer(Severity))]

expected_counties <- data.table(
  County = c("Nassau", "Suffolk", "Queens"),
  expected_n = c(19390L, 13838L, 147L)
)

expected_severity <- data.table(
  Severity = c(1L, 2L, 3L, 4L),
  expected_n = c(161L, 28357L, 3938L, 919L)
)

county_check <- merge(expected_counties, county_profile, by = "County", all.x = TRUE)
severity_check <- merge(
  expected_severity,
  severity_profile[, Severity := as.integer(Severity)],
  by = "Severity",
  all.x = TRUE
)

if (any(county_check$expected_n != county_check$N)) {
  stop("County counts do not match the execution plan expectations.")
}

if (any(severity_check$expected_n != severity_check$N)) {
  stop("Severity counts do not match the execution plan expectations.")
}

fwrite(county_profile, file.path(output_dir, "phase2_county_profile.csv"))
fwrite(severity_profile, file.path(output_dir, "phase2_severity_profile.csv"))

knitr::kable(county_check, caption = "County profile check")
knitr::kable(severity_check, caption = "Severity profile check")
```

```{r phase-2-missingness}
missingness_profile <- data.table(
  column = names(df_raw),
  missing_n = vapply(df_raw, function(column) sum(is.na(column)), integer(1)),
  missing_pct = round(vapply(df_raw, function(column) mean(is.na(column)) * 100, numeric(1)), 2)
)[order(-missing_pct, column)]

missingness_focus <- missingness_profile[
  column %in% c("End_Lat", "End_Lng", "Precipitation(in)", "Wind_Chill(F)")
]

fwrite(missingness_profile, file.path(output_dir, "phase2_missingness_profile.csv"))

knitr::kable(head(missingness_profile, 12), caption = "Top missingness profile")
knitr::kable(missingness_focus, caption = "Critical missingness checks")
```

```{r phase-2-binary-flags}
binary_flag_profile <- data.table(
  flag = c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit"),
  true_count = vapply(
    df_raw[, c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit")],
    function(column) sum(as.logical(column), na.rm = TRUE),
    integer(1)
  )
)

binary_flag_profile[, true_pct := round(100 * true_count / nrow(df_raw), 4)]

fwrite(binary_flag_profile, file.path(output_dir, "phase2_binary_flag_profile.csv"))

knitr::kable(binary_flag_profile, caption = "Near-zero binary flag checks")
```

```{r phase-2-duration-distance}
duration_values <- as.numeric(df_raw$Duration_Minutes)
distance_values <- as.numeric(df_raw[["Distance(mi)"]])

duration_profile <- data.table(
  metric = c(
    "Median duration",
    "Mean duration",
    "Max duration",
    "Rows above 1 day",
    "Rows above 1 week",
    "99th percentile"
  ),
  value = c(
    median(duration_values, na.rm = TRUE),
    mean(duration_values, na.rm = TRUE),
    max(duration_values, na.rm = TRUE),
    sum(duration_values > 1440, na.rm = TRUE),
    sum(duration_values > 10080, na.rm = TRUE),
    as.numeric(quantile(duration_values, 0.99, na.rm = TRUE))
  )
)

distance_profile <- data.table(
  metric = c("Zero-distance rows", "Zero-distance pct", "Max distance"),
  value = c(
    sum(distance_values == 0, na.rm = TRUE),
    100 * mean(distance_values == 0, na.rm = TRUE),
    max(distance_values, na.rm = TRUE)
  )
)

fwrite(duration_profile, file.path(output_dir, "phase2_duration_profile.csv"))
fwrite(distance_profile, file.path(output_dir, "phase2_distance_profile.csv"))

knitr::kable(duration_profile, caption = "Duration anomaly profile")
knitr::kable(distance_profile, caption = "Distance profile")
```

```{r phase-2-wind-and-weather}
wind_direction_profile <- as.data.table(df_raw)[, .N, by = Wind_Direction][order(-N, Wind_Direction)]
weather_profile <- as.data.table(df_raw)[, .N, by = Weather_Condition][order(-N, Weather_Condition)]

wind_focus <- wind_direction_profile[
  Wind_Direction %in% c("N", "North", "S", "South", "E", "East", "W", "West")
]

fwrite(wind_direction_profile, file.path(output_dir, "phase2_wind_direction_profile.csv"))
fwrite(weather_profile, file.path(output_dir, "phase2_weather_profile.csv"))

knitr::kable(wind_focus, caption = "Wind direction duplicate label check")
knitr::kable(head(weather_profile, 12), caption = "Top weather labels")
```

```{r phase-2-summary}
profiling_summary <- data.table(
  finding = c(
    "Dimensions match the execution plan",
    "County counts match the Long Island scope",
    "Severity counts match the execution plan",
    "End_Lat and End_Lng remain heavily missing",
    "Precipitation and Wind_Chill remain too sparse for core clustering use",
    "Five binary roadway flags carry no signal",
    "Duration still shows extreme right-tail outliers",
    "Distance includes many zero values",
    "Wind direction labels need normalization",
    "Weather labels still need grouping before clustering"
  ),
  status = c(
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed"
  )
)

fwrite(profiling_summary, file.path(output_dir, "phase2_profiling_summary.csv"))

knitr::kable(profiling_summary, caption = "Phase 2 profiling summary")
```

# Phase 3: Feature Engineering

This phase derives the analysis-ready fields needed for the clustering inputs while keeping the original cleaned dataset unchanged in `df_raw`.

```{r phase-3-start}
df_phase3 <- as.data.table(copy(df_raw))
```

```{r phase-3-temporal-features}
df_phase3[, Start_Time_Parsed := lubridate::ymd_hms(Start_Time, quiet = TRUE)]

df_phase3[, hour := lubridate::hour(Start_Time_Parsed)]
df_phase3[, month := lubridate::month(Start_Time_Parsed)]
df_phase3[, weekday := factor(
  weekdays(as.Date(Start_Time_Parsed)),
  levels = ordered_weekdays,
  ordered = TRUE
)]
df_phase3[, weekend := weekdays(as.Date(Start_Time_Parsed)) %in% c("Saturday", "Sunday")]
df_phase3[, hour_band := dplyr::case_when(
  hour >= 0 & hour <= 5 ~ "Early_AM",
  hour >= 6 & hour <= 9 ~ "AM_Rush",
  hour >= 10 & hour <= 14 ~ "Midday",
  hour >= 15 & hour <= 18 ~ "PM_Rush",
  hour >= 19 & hour <= 21 ~ "Evening",
  hour >= 22 & hour <= 23 ~ "Late_Night",
  TRUE ~ NA_character_
)]
df_phase3[, hour_band := factor(
  hour_band,
  levels = c("Early_AM", "AM_Rush", "Midday", "PM_Rush", "Evening", "Late_Night"),
  ordered = TRUE
)]
df_phase3[, season := dplyr::case_when(
  month %in% c(12L, 1L, 2L) ~ "Winter",
  month %in% c(3L, 4L, 5L) ~ "Spring",
  month %in% c(6L, 7L, 8L) ~ "Summer",
  month %in% c(9L, 10L, 11L) ~ "Fall",
  TRUE ~ NA_character_
)]
df_phase3[, season := factor(
  season,
  levels = c("Winter", "Spring", "Summer", "Fall"),
  ordered = TRUE
)]
df_phase3[, rush_hour := hour %in% c(6L, 7L, 8L, 9L, 15L, 16L, 17L, 18L)]

temporal_na_check <- data.table(
  feature = c("Start_Time_Parsed", "hour", "month", "weekday", "weekend", "hour_band", "season", "rush_hour"),
  missing_n = c(
    sum(is.na(df_phase3$Start_Time_Parsed)),
    sum(is.na(df_phase3$hour)),
    sum(is.na(df_phase3$month)),
    sum(is.na(df_phase3$weekday)),
    sum(is.na(df_phase3$weekend)),
    sum(is.na(df_phase3$hour_band)),
    sum(is.na(df_phase3$season)),
    sum(is.na(df_phase3$rush_hour))
  )
)

if (any(temporal_na_check$missing_n > 0)) {
  stop("Phase 3 temporal feature derivation created unexpected missing values.")
}

fwrite(temporal_na_check, file.path(output_dir, "phase3_temporal_check.csv"))

knitr::kable(temporal_na_check, caption = "Phase 3 temporal feature check")
```

```{r phase-3-duration}
duration_values <- as.numeric(df_phase3$Duration_Minutes)
duration_cap_99 <- as.numeric(quantile(duration_values, 0.99, na.rm = TRUE))
df_phase3[, Duration_cap := pmin(duration_values, duration_cap_99)]
df_phase3[, log_Duration := log1p(Duration_cap)]

duration_engineering_summary <- data.table(
  metric = c("99th percentile cap", "Median log_Duration", "Max log_Duration"),
  value = c(
    duration_cap_99,
    median(df_phase3$log_Duration, na.rm = TRUE),
    max(df_phase3$log_Duration, na.rm = TRUE)
  )
)

fwrite(duration_engineering_summary, file.path(output_dir, "phase3_duration_engineering.csv"))

knitr::kable(duration_engineering_summary, caption = "Phase 3 duration treatment")
```

```{r phase-3-weather-group}
weather_counts <- as.data.table(df_phase3)[, .N, by = Weather_Condition]
rare_weather <- weather_counts[N < 30, Weather_Condition]

df_phase3[, weather_group := dplyr::case_when(
  Weather_Condition %in% weather_group_map$Clear ~ "Clear",
  Weather_Condition %in% weather_group_map$Cloudy ~ "Cloudy",
  Weather_Condition %in% weather_group_map$Rain ~ "Rain",
  Weather_Condition %in% weather_group_map$Snow_Ice ~ "Snow_Ice",
  Weather_Condition %in% weather_group_map$Fog_Haze ~ "Fog_Haze",
  grepl("Thunder", Weather_Condition, ignore.case = TRUE) ~ "Thunderstorm",
  Weather_Condition %in% rare_weather ~ "Other",
  TRUE ~ "Other"
)]

df_phase3[, weather_group := factor(
  weather_group,
  levels = c("Clear", "Cloudy", "Rain", "Snow_Ice", "Fog_Haze", "Thunderstorm", "Other")
)]

weather_group_profile <- as.data.table(df_phase3)[, .N, by = weather_group][order(-N)]
fwrite(weather_group_profile, file.path(output_dir, "phase3_weather_group_profile.csv"))

knitr::kable(weather_group_profile, caption = "Phase 3 weather group profile")
```

```{r phase-3-wind-direction}
df_phase3[, wind_dir_norm := normalize_wind_direction(Wind_Direction)]

wind_direction_summary <- as.data.table(df_phase3)[, .N, by = wind_dir_norm][order(-N)]

if (length(levels(df_phase3$wind_dir_norm)) > 9) {
  stop("Phase 3 wind direction normalization produced too many levels.")
}

fwrite(wind_direction_summary, file.path(output_dir, "phase3_wind_direction_normalized.csv"))

knitr::kable(wind_direction_summary, caption = "Phase 3 normalized wind direction profile")
```

```{r phase-3-binary-audit}
zero_variance_flags <- c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit")
near_zero_flags <- c("Give_Way", "Station", "Railway", "Amenity")
retained_flags <- c("Crossing", "Junction", "Traffic_Signal", "Stop")

binary_audit <- data.table(
  flag = c(zero_variance_flags, near_zero_flags, retained_flags),
  true_count = vapply(
    df_phase3[, c(zero_variance_flags, near_zero_flags, retained_flags), with = FALSE],
    function(column) sum(as.logical(column), na.rm = TRUE),
    integer(1)
  )
)

binary_audit[, true_pct := round(100 * true_count / nrow(df_phase3), 4)]
binary_audit[, decision := c(
  rep("Exclude: zero variance", length(zero_variance_flags)),
  rep("Exclude: near-zero variance", length(near_zero_flags)),
  rep("Retain for scenario matrix", length(retained_flags))
)]

for (flag_name in retained_flags) {
  df_phase3[, (flag_name) := factor(
    as.logical(get(flag_name)),
    levels = c(FALSE, TRUE),
    labels = c("FALSE", "TRUE")
  )]
}

fwrite(binary_audit, file.path(output_dir, "phase3_binary_audit.csv"))

knitr::kable(binary_audit, caption = "Phase 3 binary flag audit")
```

```{r phase-3-summary}
phase_3_summary <- data.table(
  feature_group = c(
    "Temporal features",
    "Duration treatment",
    "Weather grouping",
    "Wind normalization",
    "Binary flag audit"
  ),
  status = c(
    "Added hour, hour_band, weekday, weekend, month, season, rush_hour",
    "Added Duration_cap and log_Duration",
    "Added weather_group with 7 levels",
    "Added wind_dir_norm with at most 9 levels",
    "Marked zero and near-zero flags for exclusion; retained core road-context flags"
  )
)

fwrite(phase_3_summary, file.path(output_dir, "phase3_summary.csv"))

knitr::kable(phase_3_summary, caption = "Phase 3 feature engineering summary")
```

# Notes for the next phases

- Phase 2 should stay read-only and focus on profiling the cleaned dataset.
- Phase 3 now derives the main engineered features needed for the scenario and spatial matrices.
- Phase 3 and later should write tables, figures, and exports into the `output/` folder.
- No `.RData` workspace files are needed for this workflow.

```{r session-info}
sessionInfo()
```
