---
title: "Long Island Clustering Workflow"
author: "AMS 597 Project Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Phase 1: Environment and Reproducibility Setup

This file is the main reproducible workflow for the Long Island clustering analysis. It is meant to run from a clean R session and use `li_accidents_clean.csv` as the only project dataset input.

```{r phase-1-config}
required_packages <- c(
  "data.table",
  "dplyr",
  "lubridate",
  "clustMixType",
  "dbscan",
  "cluster",
  "fpc",
  "ggplot2",
  "geosphere",
  "factoextra",
  "knitr",
  "rmarkdown"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste(
      "Install these packages before running the full workflow:",
      paste(missing_packages, collapse = ", ")
    )
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

set.seed(42)

input_path <- "../li_accidents_clean.csv"
output_dir <- "output"

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

if (!file.exists(input_path)) {
  stop("Expected input file not found: ", input_path)
}

phase_1_check <- data.frame(
  item = c(
    "Input dataset",
    "Output directory",
    "Seed",
    "Working directory"
  ),
  value = c(
    normalizePath(input_path, winslash = "/"),
    normalizePath(output_dir, winslash = "/"),
    "42",
    normalizePath(".", winslash = "/")
  )
)

knitr::kable(phase_1_check, caption = "Phase 1 setup check")
```

```{r load-data-preview}
df_raw <- data.table::fread(input_path, showProgress = FALSE)

data_overview <- data.frame(
  rows = nrow(df_raw),
  columns = ncol(df_raw),
  source_file = basename(input_path)
)

knitr::kable(data_overview, caption = "Input file loaded successfully")
```

# Notes for the next phases

- Phase 2 should stay read-only and focus on profiling the cleaned dataset.
- Phase 3 and later should write tables, figures, and exports into the `output/` folder.
- No `.RData` workspace files are needed for this workflow.

```{r session-info}
sessionInfo()
```
