---
title: "Long Island Clustering Workflow v2.0"
subtitle: "Corrected and Strengthened --- Based on Audit Report"
author: "AMS 597 Project Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r helpers, include=FALSE}
ordered_weekdays <- c(
  "Monday", "Tuesday", "Wednesday", "Thursday",
  "Friday", "Saturday", "Sunday"
)

weather_group_map <- list(
  Clear = c("Fair", "Clear", "Fair / Windy"),
  Cloudy = c(
    "Mostly Cloudy", "Partly Cloudy", "Scattered Clouds", "Overcast", "Cloudy",
    "Mostly Cloudy / Windy", "Partly Cloudy / Windy", "Cloudy / Windy"
  ),
  Rain = c(
    "Light Rain", "Rain", "Heavy Rain", "Light Drizzle", "Light Rain / Windy",
    "Heavy Rain / Windy", "Drizzle"
  ),
  Snow_Ice = c(
    "Light Snow", "Snow", "Heavy Snow", "Wintry Mix",
    "Light Freezing Rain", "Ice Pellets", "Sleet"
  ),
  Fog_Haze = c("Fog", "Haze", "Smoke", "Mist", "Patches of Fog", "Shallow Fog"),
  # v2.0: Thunderstorm moved from grepl fallback to explicit lookup
  Thunderstorm = c(
    "Thunderstorm", "Thunder", "Thunder / Windy",
    "Thunderstorm / Windy", "Heavy Thunderstorm",
    "Light Thunderstorm", "Thunderstorms and Rain"
  )
)

normalize_wind_direction <- function(x) {
  x <- trimws(as.character(x))
  x_upper <- toupper(x)

  mapped <- dplyr::case_when(
    x_upper %in% c("N", "NORTH") ~ "N",
    x_upper %in% c("NE", "NORTHEAST", "NORTH EAST") ~ "NE",
    x_upper %in% c("E", "EAST") ~ "E",
    x_upper %in% c("SE", "SOUTHEAST", "SOUTH EAST") ~ "SE",
    x_upper %in% c("S", "SOUTH") ~ "S",
    x_upper %in% c("SW", "SOUTHWEST", "SOUTH WEST") ~ "SW",
    x_upper %in% c("W", "WEST") ~ "W",
    x_upper %in% c("NW", "NORTHWEST", "NORTH WEST") ~ "NW",
    x_upper %in% c("CALM", "VARIABLE", "VAR") ~ "Variable_Calm",
    TRUE ~ "Variable_Calm"
  )

  factor(
    mapped,
    levels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "Variable_Calm")
  )
}

mode_value <- function(x) {
  values <- x[!is.na(x)]
  if (!length(values)) {
    return(NA_character_)
  }

  names(sort(table(values), decreasing = TRUE))[1]
}

dominant_share <- function(x) {
  values <- x[!is.na(x)]
  if (!length(values)) {
    return(NA_real_)
  }

  max(prop.table(table(values)))
}

wrap_label <- function(x, width = 24) {
  vapply(
    as.character(x),
    function(value) paste(strwrap(value, width = width), collapse = "\n"),
    character(1)
  )
}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!is.finite(rng[1]) || !is.finite(rng[2]) || diff(rng) == 0) {
    return(rep(0.5, length(x)))
  }

  (x - rng[1]) / diff(rng)
}
```

# Phase 1: Environment and Reproducibility Setup

This file is the main reproducible workflow for the Long Island clustering analysis. It is meant to run from a clean R session and use `li_accidents_clean.csv` as the only project dataset input.

```{r phase-1-config}
required_packages <- c(
  "data.table",
  "dplyr",
  "lubridate",
  "clustMixType",
  "dbscan",
  "cluster",
  "fpc",
  "ggplot2",
  "ggrepel",
  "geosphere",
  "factoextra",
  "knitr",
  "rmarkdown"
)

optional_packages <- c("plotly", "htmlwidgets")

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste(
      "Install these packages before running the full workflow:",
      paste(missing_packages, collapse = ", ")
    )
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))

set.seed(42)

input_path <- "../li_accidents_clean.csv"
output_dir <- "local_artifacts/output"
run_profile <- "consensus_16gb"

if (!run_profile %in% c("local_safe", "consensus_16gb", "full_srs")) {
  stop("run_profile must be one of 'local_safe', 'consensus_16gb', or 'full_srs'.")
}

phase6_validation_mode <- dplyr::case_when(
  run_profile == "full_srs" ~ "full",
  run_profile == "consensus_16gb" ~ "consensus_sampled",
  TRUE ~ "sampled"
)
phase6_validation_sample_n <- dplyr::case_when(
  run_profile == "full_srs" ~ NA_integer_,
  run_profile == "consensus_16gb" ~ 4000L,
  TRUE ~ 5000L
)
phase6_validation_repeats <- dplyr::case_when(
  run_profile == "consensus_16gb" ~ 4L,
  TRUE ~ 1L
)

phase6_stability_mode <- dplyr::case_when(
  run_profile == "full_srs" ~ "full",
  run_profile == "consensus_16gb" ~ "consensus_sampled",
  TRUE ~ "sampled"
)
phase6_stability_sample_n <- dplyr::case_when(
  run_profile == "full_srs" ~ NA_integer_,
  run_profile == "consensus_16gb" ~ 2500L,
  TRUE ~ 3000L
)
phase6_stability_repeats <- dplyr::case_when(
  run_profile == "consensus_16gb" ~ 3L,
  TRUE ~ 1L
)
phase6_stability_bootstraps <- dplyr::case_when(
  run_profile == "full_srs" ~ 50L,
  run_profile == "consensus_16gb" ~ 10L,
  TRUE ~ 15L
)
phase9_interactive_mode <- "auto"

interactive_exports_enabled <- dplyr::case_when(
  phase9_interactive_mode == "off" ~ FALSE,
  phase9_interactive_mode == "on" ~ all(
    vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)
  ),
  phase9_interactive_mode == "auto" ~ all(
    vapply(optional_packages, requireNamespace, logical(1), quietly = TRUE)
  ),
  TRUE ~ FALSE
)

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

if (!file.exists(input_path)) {
  stop("Expected input file not found: ", input_path)
}

phase_1_check <- data.frame(
  item = c(
    "Input dataset",
    "Output directory",
    "Seed",
    "Run profile",
    "Phase 6 validation mode",
    "Phase 6 sample size",
    "Phase 6 validation repeats",
    "Phase 6 stability mode",
    "Phase 6 stability sample size",
    "Phase 6 stability repeats",
    "Phase 6 stability bootstraps",
    "Phase 9 interactive mode",
    "Working directory"
  ),
  value = c(
    normalizePath(input_path, winslash = "/"),
    normalizePath(output_dir, winslash = "/"),
    "42",
    run_profile,
    phase6_validation_mode,
    ifelse(is.na(phase6_validation_sample_n), "Full data", as.character(phase6_validation_sample_n)),
    as.character(phase6_validation_repeats),
    phase6_stability_mode,
    ifelse(is.na(phase6_stability_sample_n), "Full data", as.character(phase6_stability_sample_n)),
    as.character(phase6_stability_repeats),
    as.character(phase6_stability_bootstraps),
    ifelse(
      interactive_exports_enabled,
      "Enabled",
      ifelse(phase9_interactive_mode == "off", "Off for safe local runs", "Packages missing")
    ),
    normalizePath(".", winslash = "/")
  )
)

knitr::kable(phase_1_check, caption = "Phase 1 setup check")
```

```{r load-data-preview}
df_raw <- data.table::fread(input_path, showProgress = FALSE)

data_overview <- data.frame(
  rows = nrow(df_raw),
  columns = ncol(df_raw),
  source_file = basename(input_path)
)

knitr::kable(data_overview, caption = "Input file loaded successfully")
```

# Phase 2: Data Ingestion and Raw Profiling

This phase stays read-only. The goal is to confirm that the cleaned Long Island dataset matches the expectations in the execution plan before any feature engineering starts.

```{r phase-2-dimension-checks}
expected_rows <- 33375L
expected_cols <- 48L

if (nrow(df_raw) != expected_rows || ncol(df_raw) != expected_cols) {
  stop(
    paste0(
      "Unexpected dataset dimensions. Expected ",
      expected_rows, " rows and ", expected_cols, " columns, got ",
      nrow(df_raw), " rows and ", ncol(df_raw), " columns."
    )
  )
}

phase_2_dimensions <- data.frame(
  metric = c("Rows", "Columns"),
  expected = c(expected_rows, expected_cols),
  observed = c(nrow(df_raw), ncol(df_raw))
)

knitr::kable(phase_2_dimensions, caption = "Phase 2 dimension check")
```

```{r phase-2-county-and-severity}
county_profile <- as.data.table(df_raw)[, .N, by = County][order(match(County, c("Nassau", "Suffolk", "Queens")))]
severity_profile <- as.data.table(df_raw)[, .N, by = Severity][order(as.integer(Severity))]

expected_counties <- data.table(
  County = c("Nassau", "Suffolk", "Queens"),
  expected_n = c(19390L, 13838L, 147L)
)

expected_severity <- data.table(
  Severity = c(1L, 2L, 3L, 4L),
  expected_n = c(161L, 28357L, 3938L, 919L)
)

county_check <- merge(expected_counties, county_profile, by = "County", all.x = TRUE)
severity_check <- merge(
  expected_severity,
  severity_profile[, Severity := as.integer(Severity)],
  by = "Severity",
  all.x = TRUE
)

if (any(county_check$expected_n != county_check$N)) {
  stop("County counts do not match the execution plan expectations.")
}

if (any(severity_check$expected_n != severity_check$N)) {
  stop("Severity counts do not match the execution plan expectations.")
}

fwrite(county_profile, file.path(output_dir, "phase2_county_profile.csv"))
fwrite(severity_profile, file.path(output_dir, "phase2_severity_profile.csv"))

knitr::kable(county_check, caption = "County profile check")
knitr::kable(severity_check, caption = "Severity profile check")
```

```{r phase-2-missingness}
missingness_profile <- data.table(
  column = names(df_raw),
  missing_n = vapply(df_raw, function(column) sum(is.na(column)), integer(1)),
  missing_pct = round(vapply(df_raw, function(column) mean(is.na(column)) * 100, numeric(1)), 2)
)[order(-missing_pct, column)]

missingness_focus <- missingness_profile[
  column %in% c("End_Lat", "End_Lng", "Precipitation(in)", "Wind_Chill(F)")
]

fwrite(missingness_profile, file.path(output_dir, "phase2_missingness_profile.csv"))

knitr::kable(head(missingness_profile, 12), caption = "Top missingness profile")
knitr::kable(missingness_focus, caption = "Critical missingness checks")
```

```{r phase-2-binary-flags}
binary_flag_profile <- data.table(
  flag = c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit"),
  true_count = vapply(
    df_raw[, c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit")],
    function(column) sum(as.logical(column), na.rm = TRUE),
    integer(1)
  )
)

binary_flag_profile[, true_pct := round(100 * true_count / nrow(df_raw), 4)]

fwrite(binary_flag_profile, file.path(output_dir, "phase2_binary_flag_profile.csv"))

knitr::kable(binary_flag_profile, caption = "Near-zero binary flag checks")
```

```{r phase-2-duration-distance}
duration_values <- as.numeric(df_raw$Duration_Minutes)
distance_values <- as.numeric(df_raw[["Distance(mi)"]])

duration_profile <- data.table(
  metric = c(
    "Median duration",
    "Mean duration",
    "Max duration",
    "Rows above 1 day",
    "Rows above 1 week",
    "99th percentile"
  ),
  value = c(
    median(duration_values, na.rm = TRUE),
    mean(duration_values, na.rm = TRUE),
    max(duration_values, na.rm = TRUE),
    sum(duration_values > 1440, na.rm = TRUE),
    sum(duration_values > 10080, na.rm = TRUE),
    as.numeric(quantile(duration_values, 0.99, na.rm = TRUE))
  )
)

distance_profile <- data.table(
  metric = c("Zero-distance rows", "Zero-distance pct", "Max distance"),
  value = c(
    sum(distance_values == 0, na.rm = TRUE),
    100 * mean(distance_values == 0, na.rm = TRUE),
    max(distance_values, na.rm = TRUE)
  )
)

fwrite(duration_profile, file.path(output_dir, "phase2_duration_profile.csv"))
fwrite(distance_profile, file.path(output_dir, "phase2_distance_profile.csv"))

knitr::kable(duration_profile, caption = "Duration anomaly profile")
knitr::kable(distance_profile, caption = "Distance profile")
```

```{r phase-2-wind-and-weather}
wind_direction_profile <- as.data.table(df_raw)[, .N, by = Wind_Direction][order(-N, Wind_Direction)]
weather_profile <- as.data.table(df_raw)[, .N, by = Weather_Condition][order(-N, Weather_Condition)]

wind_focus <- wind_direction_profile[
  Wind_Direction %in% c("N", "North", "S", "South", "E", "East", "W", "West")
]

fwrite(wind_direction_profile, file.path(output_dir, "phase2_wind_direction_profile.csv"))
fwrite(weather_profile, file.path(output_dir, "phase2_weather_profile.csv"))

knitr::kable(wind_focus, caption = "Wind direction duplicate label check")
knitr::kable(head(weather_profile, 12), caption = "Top weather labels")
```

```{r phase-2-summary}
profiling_summary <- data.table(
  finding = c(
    "Dimensions match the execution plan",
    "County counts match the Long Island scope",
    "Severity counts match the execution plan",
    "End_Lat and End_Lng remain heavily missing",
    "Precipitation and Wind_Chill remain too sparse for core clustering use",
    "Five binary roadway flags carry no signal",
    "Duration still shows extreme right-tail outliers",
    "Distance includes many zero values",
    "Wind direction labels need normalization",
    "Weather labels still need grouping before clustering"
  ),
  status = c(
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed",
    "Confirmed"
  )
)

fwrite(profiling_summary, file.path(output_dir, "phase2_profiling_summary.csv"))

knitr::kable(profiling_summary, caption = "Phase 2 profiling summary")
```

# Phase 3: Feature Engineering

This phase derives the analysis-ready fields needed for the clustering inputs while keeping the original cleaned dataset unchanged in `df_raw`.

```{r phase-3-start}
df_phase3 <- as.data.table(copy(df_raw))
```

```{r phase-3-temporal-features}
df_phase3[, Start_Time_Parsed := lubridate::ymd_hms(Start_Time, quiet = TRUE)]

df_phase3[, hour := lubridate::hour(Start_Time_Parsed)]
df_phase3[, month := lubridate::month(Start_Time_Parsed)]
df_phase3[, weekday := factor(
  weekdays(as.Date(Start_Time_Parsed)),
  levels = ordered_weekdays,
  ordered = TRUE
)]
df_phase3[, weekend := weekdays(as.Date(Start_Time_Parsed)) %in% c("Saturday", "Sunday")]
df_phase3[, hour_band := dplyr::case_when(
  hour >= 0 & hour <= 5 ~ "Early_AM",
  hour >= 6 & hour <= 9 ~ "AM_Rush",
  hour >= 10 & hour <= 14 ~ "Midday",
  hour >= 15 & hour <= 18 ~ "PM_Rush",
  hour >= 19 & hour <= 21 ~ "Evening",
  hour >= 22 & hour <= 23 ~ "Late_Night",
  TRUE ~ NA_character_
)]
df_phase3[, hour_band := factor(
  hour_band,
  levels = c("Early_AM", "AM_Rush", "Midday", "PM_Rush", "Evening", "Late_Night"),
  ordered = TRUE
)]
df_phase3[, season := dplyr::case_when(
  month %in% c(12L, 1L, 2L) ~ "Winter",
  month %in% c(3L, 4L, 5L) ~ "Spring",
  month %in% c(6L, 7L, 8L) ~ "Summer",
  month %in% c(9L, 10L, 11L) ~ "Fall",
  TRUE ~ NA_character_
)]
df_phase3[, season := factor(
  season,
  levels = c("Winter", "Spring", "Summer", "Fall"),
  ordered = TRUE
)]
df_phase3[, rush_hour := hour %in% c(6L, 7L, 8L, 9L, 15L, 16L, 17L, 18L)]

temporal_na_check <- data.table(
  feature = c("Start_Time_Parsed", "hour", "month", "weekday", "weekend", "hour_band", "season", "rush_hour"),
  missing_n = c(
    sum(is.na(df_phase3$Start_Time_Parsed)),
    sum(is.na(df_phase3$hour)),
    sum(is.na(df_phase3$month)),
    sum(is.na(df_phase3$weekday)),
    sum(is.na(df_phase3$weekend)),
    sum(is.na(df_phase3$hour_band)),
    sum(is.na(df_phase3$season)),
    sum(is.na(df_phase3$rush_hour))
  )
)

if (any(temporal_na_check$missing_n > 0)) {
  stop("Phase 3 temporal feature derivation created unexpected missing values.")
}

fwrite(temporal_na_check, file.path(output_dir, "phase3_temporal_check.csv"))

knitr::kable(temporal_na_check, caption = "Phase 3 temporal feature check")
```

```{r phase-3-duration}
duration_values <- as.numeric(df_phase3$Duration_Minutes)
duration_cap_99 <- as.numeric(quantile(duration_values, 0.99, na.rm = TRUE))
df_phase3[, Duration_cap := pmin(duration_values, duration_cap_99)]
df_phase3[, log_Duration := log1p(Duration_cap)]

duration_engineering_summary <- data.table(
  metric = c("99th percentile cap", "Median log_Duration", "Max log_Duration"),
  value = c(
    duration_cap_99,
    median(df_phase3$log_Duration, na.rm = TRUE),
    max(df_phase3$log_Duration, na.rm = TRUE)
  )
)

fwrite(duration_engineering_summary, file.path(output_dir, "phase3_duration_engineering.csv"))

knitr::kable(duration_engineering_summary, caption = "Phase 3 duration treatment")
```

```{r phase-3-weather-group}
weather_counts <- as.data.table(df_phase3)[, .N, by = Weather_Condition]
rare_weather <- weather_counts[N < 30, Weather_Condition]

df_phase3[, weather_group := dplyr::case_when(
  Weather_Condition %in% weather_group_map$Clear       ~ "Clear",
  Weather_Condition %in% weather_group_map$Cloudy      ~ "Cloudy",
  Weather_Condition %in% weather_group_map$Rain        ~ "Rain",
  Weather_Condition %in% weather_group_map$Snow_Ice    ~ "Snow_Ice",
  Weather_Condition %in% weather_group_map$Fog_Haze    ~ "Fog_Haze",
  # v2.0: explicit lookup first; grepl catches novel variants not in the map
  Weather_Condition %in% weather_group_map$Thunderstorm ~ "Thunderstorm",
  grepl("Thunder", Weather_Condition, ignore.case = TRUE) ~ "Thunderstorm",
  Weather_Condition %in% rare_weather ~ "Other",
  TRUE ~ "Other"
)]

df_phase3[, weather_group := factor(
  weather_group,
  levels = c("Clear", "Cloudy", "Rain", "Snow_Ice", "Fog_Haze", "Thunderstorm", "Other")
)]

weather_group_profile <- as.data.table(df_phase3)[, .N, by = weather_group][order(-N)]
fwrite(weather_group_profile, file.path(output_dir, "phase3_weather_group_profile.csv"))

knitr::kable(weather_group_profile, caption = "Phase 3 weather group profile")
```

```{r phase-3-wind-direction}
df_phase3[, wind_dir_norm := normalize_wind_direction(Wind_Direction)]

wind_direction_summary <- as.data.table(df_phase3)[, .N, by = wind_dir_norm][order(-N)]

if (length(levels(df_phase3$wind_dir_norm)) > 9) {
  stop("Phase 3 wind direction normalization produced too many levels.")
}

fwrite(wind_direction_summary, file.path(output_dir, "phase3_wind_direction_normalized.csv"))

knitr::kable(wind_direction_summary, caption = "Phase 3 normalized wind direction profile")
```

```{r phase-3-binary-audit}
zero_variance_flags <- c("Bump", "Roundabout", "Turning_Loop", "Traffic_Calming", "No_Exit")
near_zero_flags <- c("Give_Way", "Station", "Railway", "Amenity")
retained_flags <- c("Crossing", "Junction", "Traffic_Signal", "Stop")

binary_audit <- data.table(
  flag = c(zero_variance_flags, near_zero_flags, retained_flags),
  true_count = vapply(
    df_phase3[, c(zero_variance_flags, near_zero_flags, retained_flags), with = FALSE],
    function(column) sum(as.logical(column), na.rm = TRUE),
    integer(1)
  )
)

binary_audit[, true_pct := round(100 * true_count / nrow(df_phase3), 4)]
binary_audit[, decision := c(
  rep("Exclude: zero variance", length(zero_variance_flags)),
  rep("Exclude: near-zero variance", length(near_zero_flags)),
  rep("Retain for scenario matrix", length(retained_flags))
)]

for (flag_name in retained_flags) {
  # v2.0: use native FALSE/TRUE factor levels (no string labels)
  df_phase3[, (flag_name) := factor(
    as.logical(get(flag_name)),
    levels = c(FALSE, TRUE)
  )]
}

fwrite(binary_audit, file.path(output_dir, "phase3_binary_audit.csv"))

knitr::kable(binary_audit, caption = "Phase 3 binary flag audit")
```

```{r phase-3-summary}
phase_3_summary <- data.table(
  feature_group = c(
    "Temporal features",
    "Duration treatment",
    "Weather grouping",
    "Wind normalization",
    "Binary flag audit"
  ),
  status = c(
    "Added hour, hour_band, weekday, weekend, month, season, rush_hour",
    "Added Duration_cap and log_Duration",
    "Added weather_group with 7 levels",
    "Added wind_dir_norm with at most 9 levels",
    "Marked zero and near-zero flags for exclusion; retained core road-context flags"
  )
)

fwrite(phase_3_summary, file.path(output_dir, "phase3_summary.csv"))

knitr::kable(phase_3_summary, caption = "Phase 3 feature engineering summary")
```

# Phase 4: Build Clustering Input Matrices

This phase constructs the final matrices that will be used by the clustering methods. Nothing outside these matrices should enter the clustering algorithms.

```{r phase-4-scenario-matrix}
df_scenario_base <- as.data.table(copy(df_phase3))

# v2.0: native FALSE/TRUE levels, no string labels
df_scenario_base[, weekend   := factor(weekend,   levels = c(FALSE, TRUE))]
df_scenario_base[, rush_hour := factor(rush_hour, levels = c(FALSE, TRUE))]
df_scenario_base[, Sunrise_Sunset := factor(Sunrise_Sunset)]

scenario_factor_cols <- c(
  "hour_band", "weekday", "weekend", "season", "rush_hour",
  "Sunrise_Sunset", "weather_group", "Crossing", "Junction",
  "Traffic_Signal", "Stop"
)

scenario_numeric_cols <- c(
  "Temperature(F)", "Humidity(%)", "Pressure(in)",
  "Visibility(mi)", "Wind_Speed(mph)"
)

if (mean(is.na(df_scenario_base$log_Duration)) < 0.05) {
  scenario_numeric_cols <- c(scenario_numeric_cols, "log_Duration")
}

scenario_columns <- c(scenario_factor_cols, scenario_numeric_cols)

df_scenario <- df_scenario_base[, ..scenario_columns]

excluded_columns <- c(
  "Severity", "ID", "Source", "Description", "Street", "City", "Zipcode", "nzip",
  "Start_Lat", "Start_Lng", "End_Lat", "End_Lng", "County", "Distance(mi)",
  "Precipitation(in)", "Wind_Chill(F)", "Bump", "Roundabout", "Turning_Loop",
  "Traffic_Calming", "No_Exit", "Give_Way", "Station", "Railway", "Amenity"
)

unexpected_excluded <- intersect(names(df_scenario), excluded_columns)
if (length(unexpected_excluded) > 0) {
  stop(
    paste(
      "Excluded columns were found in df_scenario:",
      paste(unexpected_excluded, collapse = ", ")
    )
  )
}

numeric_imputation_log <- data.table(
  column = character(),
  missing_before = integer(),
  missing_pct = numeric(),
  imputation = character()
)

for (column_name in scenario_numeric_cols) {
  missing_before <- sum(is.na(df_scenario[[column_name]]))
  if (!missing_before) {
    next
  }

  missing_pct <- 100 * mean(is.na(df_scenario[[column_name]]))
  median_value <- median(df_scenario[[column_name]], na.rm = TRUE)
  df_scenario[is.na(get(column_name)), (column_name) := median_value]

  numeric_imputation_log <- rbind(
    numeric_imputation_log,
    data.table(
      column = column_name,
      missing_before = missing_before,
      missing_pct = round(missing_pct, 4),
      imputation = "Median"
    )
  )
}

factor_imputation_log <- data.table(
  column = character(),
  missing_before = integer(),
  rule_applied = character()
)

for (column_name in scenario_factor_cols) {
  missing_before <- sum(is.na(df_scenario[[column_name]]))
  if (!missing_before) {
    next
  }

  if (missing_before < 10) {
    fill_value <- mode_value(df_scenario[[column_name]])
    df_scenario[is.na(get(column_name)), (column_name) := fill_value]
    rule_applied <- "Mode"
  } else {
    df_scenario[[column_name]] <- addNA(df_scenario[[column_name]])
    levels(df_scenario[[column_name]])[is.na(levels(df_scenario[[column_name]]))] <- "Unknown"
    rule_applied <- "Unknown level"
  }

  factor_imputation_log <- rbind(
    factor_imputation_log,
    data.table(
      column = column_name,
      missing_before = missing_before,
      rule_applied = rule_applied
    )
  )
}

scenario_na_total <- sum(is.na(df_scenario))
if (scenario_na_total > 0) {
  stop("df_scenario still contains missing values after Phase 4 imputations.")
}

scenario_column_profile <- data.table(
  column = names(df_scenario),
  type = vapply(df_scenario, function(column) class(column)[1], character(1))
)

scenario_matrix_summary <- data.table(
  metric = c("Rows", "Columns", "Factor columns", "Numeric columns", "Total NAs"),
  value = c(
    nrow(df_scenario),
    ncol(df_scenario),
    length(scenario_factor_cols),
    length(scenario_numeric_cols),
    scenario_na_total
  )
)

fwrite(scenario_column_profile, file.path(output_dir, "phase4_scenario_columns.csv"))
fwrite(numeric_imputation_log, file.path(output_dir, "phase4_numeric_imputation_log.csv"))
fwrite(factor_imputation_log, file.path(output_dir, "phase4_factor_imputation_log.csv"))
fwrite(scenario_matrix_summary, file.path(output_dir, "phase4_scenario_matrix_summary.csv"))

knitr::kable(scenario_matrix_summary, caption = "Phase 4 scenario matrix summary")
knitr::kable(scenario_column_profile, caption = "Phase 4 scenario column list")
```

```{r phase-4-spatial-matrix}
if (any(is.na(df_phase3$Start_Lat)) || any(is.na(df_phase3$Start_Lng))) {
  stop("Start_Lat or Start_Lng contains missing values, so df_spatial cannot be built.")
}

lat0 <- 40.8
lng0 <- -73.2

df_spatial <- data.table(
  x_km = (df_phase3$Start_Lng - lng0) * 111.32 * cos(lat0 * pi / 180),
  y_km = (df_phase3$Start_Lat - lat0) * 110.574
)

spatial_extent <- data.table(
  metric = c("Rows", "Columns", "x_km min", "x_km max", "x_km span", "y_km min", "y_km max", "y_km span"),
  value = c(
    nrow(df_spatial),
    ncol(df_spatial),
    min(df_spatial$x_km),
    max(df_spatial$x_km),
    diff(range(df_spatial$x_km)),
    min(df_spatial$y_km),
    max(df_spatial$y_km),
    diff(range(df_spatial$y_km))
  )
)

if (sum(is.na(df_spatial)) > 0) {
  stop("df_spatial contains missing values after coordinate projection.")
}

fwrite(spatial_extent, file.path(output_dir, "phase4_spatial_extent.csv"))

knitr::kable(spatial_extent, caption = "Phase 4 spatial matrix extent")
```

```{r phase-4-summary}
phase_4_summary <- data.table(
  matrix = c("df_scenario", "df_spatial"),
  rows = c(nrow(df_scenario), nrow(df_spatial)),
  columns = c(ncol(df_scenario), ncol(df_spatial)),
  na_count = c(sum(is.na(df_scenario)), sum(is.na(df_spatial)))
)

fwrite(phase_4_summary, file.path(output_dir, "phase4_summary.csv"))

knitr::kable(phase_4_summary, caption = "Phase 4 matrix readiness summary")
```

```{r phase-4-reduced-scenario-matrix}
# v2.0: Reduced feature matrix for sensitivity analysis.
# Drops Pressure(in) and Visibility(mi) — both required median imputation.
# Used in Phase 5 and Phase 6 to test whether removing imputed features
# improves silhouette and Jaccard scores.

reduced_numeric_cols <- setdiff(
  scenario_numeric_cols,
  c("Pressure(in)", "Visibility(mi)")
)

df_scenario_reduced <- df_scenario[
  , c(scenario_factor_cols, reduced_numeric_cols), with = FALSE
]

stopifnot(sum(is.na(df_scenario_reduced)) == 0)
stopifnot(
  ncol(df_scenario_reduced) ==
    length(scenario_factor_cols) + length(reduced_numeric_cols)
)

reduced_matrix_summary <- data.table(
  item = c(
    "Rows",
    "Factor columns",
    "Numeric columns",
    "Dropped from full matrix"
  ),
  value = c(
    nrow(df_scenario_reduced),
    length(scenario_factor_cols),
    length(reduced_numeric_cols),
    paste(c("Pressure(in)", "Visibility(mi)"), collapse = ", ")
  )
)

fwrite(
  reduced_matrix_summary,
  file.path(output_dir, "phase4_reduced_matrix_summary.csv")
)

knitr::kable(
  reduced_matrix_summary,
  caption = "Phase 4 reduced scenario matrix (dropped imputed features)"
)
```

# Phase 5: k-Prototypes Scenario Clustering

This phase runs the candidate mixed-type clustering solutions on `df_scenario` and stores the diagnostics needed for later model selection.

```{r phase-5-type-check}
scenario_type_check <- data.table(
  column = names(df_scenario),
  class = vapply(df_scenario, function(column) class(column)[1], character(1))
)

factor_check <- scenario_type_check[column %in% scenario_factor_cols]
numeric_check <- scenario_type_check[column %in% scenario_numeric_cols]

if (!all(vapply(df_scenario[, ..scenario_factor_cols], is.factor, logical(1)))) {
  stop("One or more Phase 5 scenario factor columns are not stored as factors.")
}

if (!all(vapply(df_scenario[, ..scenario_numeric_cols], is.numeric, logical(1)))) {
  stop("One or more Phase 5 scenario numeric columns are not stored as numeric values.")
}

fwrite(scenario_type_check, file.path(output_dir, "phase5_scenario_type_check.csv"))

knitr::kable(scenario_type_check, caption = "Phase 5 scenario type check")
```

```{r phase-5-kproto-candidates}
candidate_k <- 3:8
kp_results <- list()

for (k_value in candidate_k) {
  set.seed(42)
  kp_results[[as.character(k_value)]] <- clustMixType::kproto(
    x = df_scenario,
    k = k_value,
    lambda = NULL,
    nstart = 10,
    type = "gower",
    keep.data = TRUE,
    verbose = FALSE
  )
}
```

```{r phase-5-cluster-summaries}
cluster_fit_summary <- rbindlist(
  lapply(names(kp_results), function(k_label) {
    model_fit <- kp_results[[k_label]]
    cluster_sizes <- table(model_fit$cluster)

    data.table(
      k = as.integer(k_label),
      tot_withinss = model_fit$tot.withinss,
      lambda = if (is.null(model_fit$lambda)) NA_real_ else as.numeric(model_fit$lambda)[1],
      min_cluster_size = min(cluster_sizes),
      max_cluster_size = max(cluster_sizes),
      n_clusters = length(cluster_sizes),
      potentially_unstable = min(cluster_sizes) < 200
    )
  })
)

# v2.0: Add approximate numeric R² (variance explained) to fit summary.
# Note: tot.withinss in kproto includes both numeric and categorical
# components. This R² is computed on the numeric columns only and should
# be labelled "approximate" in any report table.
total_ss_numeric <- sum(
  sapply(scenario_numeric_cols, function(col) {
    x <- df_scenario[[col]]
    sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE)
  })
)

cluster_fit_summary[, variance_explained_pct := round(
  100 * (1 - tot_withinss / total_ss_numeric), 2
)]

cluster_size_table <- rbindlist(
  lapply(names(kp_results), function(k_label) {
    model_fit <- kp_results[[k_label]]
    data.table(
      k = as.integer(k_label),
      cluster = names(table(model_fit$cluster)),
      size = as.integer(table(model_fit$cluster))
    )
  })
)

dominance_checks <- rbindlist(
  lapply(names(kp_results), function(k_label) {
    model_fit <- kp_results[[k_label]]
    clustered_data <- copy(df_scenario)
    clustered_data[, cluster := factor(model_fit$cluster)]

    rbindlist(lapply(scenario_factor_cols, function(column_name) {
      clustered_data[, .(
        dominant_share = round(dominant_share(get(column_name)), 4)
      ), by = cluster][
        , `:=`(
          k = as.integer(k_label),
          feature = column_name
        )
      ]
    }), fill = TRUE)
  }),
  fill = TRUE
)

dominance_flags <- dominance_checks[
  dominant_share >= 0.9
][order(k, cluster, feature)]

fwrite(cluster_fit_summary, file.path(output_dir, "phase5_kproto_fit_summary.csv"))
fwrite(cluster_size_table, file.path(output_dir, "phase5_kproto_cluster_sizes.csv"))
fwrite(dominance_checks, file.path(output_dir, "phase5_kproto_dominance_checks.csv"))
fwrite(dominance_flags, file.path(output_dir, "phase5_kproto_dominance_flags.csv"))

knitr::kable(cluster_fit_summary, caption = "Phase 5 k-prototypes fit summary")
knitr::kable(cluster_size_table, caption = "Phase 5 cluster sizes by k")
```

```{r phase-5-elbow-plot, fig.height=4.5, fig.width=7}
elbow_plot <- ggplot2::ggplot(
  cluster_fit_summary,
  ggplot2::aes(x = k, y = tot_withinss)
) +
  ggplot2::geom_line(color = "#1f78b4", linewidth = 0.8) +
  ggplot2::geom_point(color = "#1f78b4", size = 2.4) +
  ggplot2::geom_text(
    ggplot2::aes(label = k),
    nudge_y = max(cluster_fit_summary$tot_withinss) * 0.01,
    size = 3.5
  ) +
  ggplot2::labs(
    title = "Phase 5 Elbow Plot",
    x = "Number of clusters (k)",
    y = "Total within-cluster distance"
  ) +
  ggplot2::theme_minimal()

print(elbow_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase5_kproto_elbow_plot.png"),
  plot = elbow_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

```{r phase-5-summary}
phase_5_summary <- data.table(
  checkpoint = c(
    "Candidate solutions created",
    "k range covered",
    "Elbow plot saved",
    "Cluster sizes recorded",
    "Dominance checks recorded"
  ),
  status = c(
    length(kp_results) == 6,
    paste(candidate_k, collapse = ", "),
    file.exists(file.path(output_dir, "phase5_kproto_elbow_plot.png")),
    nrow(cluster_size_table) > 0,
    nrow(dominance_checks) > 0
  )
)

fwrite(phase_5_summary, file.path(output_dir, "phase5_summary.csv"))

knitr::kable(phase_5_summary, caption = "Phase 5 readiness summary")
```

```{r phase-5-reduced-kproto}
# v2.0: Run k-prototypes on the reduced feature matrix (no Pressure/Visibility).
# Results compared against full-feature runs in Phase 6.

kp_results_reduced <- list()
for (k_val in candidate_k) {
  set.seed(42)
  kp_results_reduced[[as.character(k_val)]] <- clustMixType::kproto(
    x         = df_scenario_reduced,
    k         = k_val,
    lambda    = NULL,
    nstart    = 10,
    type      = "gower",
    keep.data = TRUE,
    verbose   = FALSE
  )
}

reduced_fit_summary <- rbindlist(
  lapply(names(kp_results_reduced), function(k_label) {
    model_fit    <- kp_results_reduced[[k_label]]
    cluster_sizes <- table(model_fit$cluster)
    data.table(
      k             = as.integer(k_label),
      tot_withinss  = model_fit$tot.withinss,
      min_cluster_size = min(cluster_sizes),
      max_cluster_size = max(cluster_sizes),
      feature_set   = "reduced"
    )
  })
)

full_fit_labeled <- copy(
  cluster_fit_summary[, .(k, tot_withinss, min_cluster_size, max_cluster_size)]
)[, feature_set := "full"]

feature_comparison_fits <- rbind(full_fit_labeled, reduced_fit_summary)
setorder(feature_comparison_fits, k, feature_set)

fwrite(
  feature_comparison_fits,
  file.path(output_dir, "phase5_feature_comparison_fits.csv")
)

knitr::kable(
  feature_comparison_fits,
  caption = "Phase 5: within-cluster distance — full vs reduced feature set"
)
```

# Phase 6: Cluster Validation and Stability

This phase uses internal validation, bootstrap stability, and cluster profile checks to choose a defensible value of `k`.

```{r phase-6-internal-validation}
validation_input <- copy(df_scenario)
validation_rows <- nrow(df_scenario)
validation_repeat_results <- NULL

if (identical(phase6_validation_mode, "consensus_sampled")) {
  validation_rows <- min(phase6_validation_sample_n, nrow(df_scenario))
  validation_repeat_results <- rbindlist(
    lapply(seq_len(phase6_validation_repeats), function(repeat_id) {
      set.seed(42 + repeat_id)
      sampled_index <- sort(sample.int(nrow(df_scenario), validation_rows))
      validation_input_repeat <- copy(df_scenario[sampled_index])

      val_repeat <- clustMixType::validation_kproto(
        method = "silhouette",
        data = validation_input_repeat,
        k = candidate_k,
        nstart = 5,
        type = "gower",
        kp_obj = "all",
        verbose = FALSE
      )

      data.table(
        repeat_id = repeat_id,
        validation_rows = validation_rows,
        repeat_k_opt = as.integer(val_repeat$k_opt),
        k = as.integer(names(val_repeat$indices)),
        silhouette = as.numeric(val_repeat$indices)
      )
    }),
    fill = TRUE
  )

  validation_optimal_counts <- unique(
    validation_repeat_results[, .(repeat_id, repeat_k_opt)]
  )[, .(optimal_repeat_count = .N), by = .(k = repeat_k_opt)]

  validation_table <- validation_repeat_results[, .(
    silhouette = mean(silhouette),
    sd_silhouette = stats::sd(silhouette),
    median_silhouette = stats::median(silhouette)
  ), by = k]

  validation_table <- merge(
    validation_table,
    validation_optimal_counts,
    by = "k",
    all.x = TRUE
  )
  validation_table[is.na(optimal_repeat_count), optimal_repeat_count := 0L]
  validation_table[, optimal_repeat_share := round(optimal_repeat_count / phase6_validation_repeats, 4)]
  validation_table <- validation_table[order(-optimal_repeat_count, -silhouette, k)]
  validation_optimal_k <- validation_table$k[1]
} else {
  if (identical(phase6_validation_mode, "sampled")) {
    set.seed(42)
    validation_rows <- min(phase6_validation_sample_n, nrow(df_scenario))
    sampled_index <- sort(sample.int(nrow(df_scenario), validation_rows))
    validation_input <- copy(df_scenario[sampled_index])
  }

  val_result <- clustMixType::validation_kproto(
    method = "silhouette",
    data = validation_input,
    k = candidate_k,
    nstart = 5,
    type = "gower",
    kp_obj = "all",
    verbose = FALSE
  )

  validation_table <- data.table(
    k = as.integer(names(val_result$indices)),
    silhouette = as.numeric(val_result$indices),
    sd_silhouette = NA_real_,
    median_silhouette = as.numeric(val_result$indices),
    optimal_repeat_count = 1L,
    optimal_repeat_share = 1
  )[order(-silhouette, k)]
  validation_optimal_k <- as.integer(val_result$k_opt)
}

validation_metadata <- data.table(
  setting = c("validation_mode", "validation_rows", "full_rows", "sample_fraction", "validation_repeats"),
  value = c(
    phase6_validation_mode,
    validation_rows,
    nrow(df_scenario),
    round(validation_rows / nrow(df_scenario), 4),
    phase6_validation_repeats
  )
)

top_two_k <- head(validation_table$k, 2)

fwrite(validation_table, file.path(output_dir, "phase6_validation_indices.csv"))
fwrite(validation_metadata, file.path(output_dir, "phase6_validation_metadata.csv"))
if (!is.null(validation_repeat_results)) {
  fwrite(validation_repeat_results, file.path(output_dir, "phase6_validation_repeats.csv"))
}

knitr::kable(validation_table, caption = "Phase 6 internal validation indices")
knitr::kable(validation_metadata, caption = "Phase 6 validation run settings")
if (!is.null(validation_repeat_results)) {
  knitr::kable(validation_repeat_results, caption = "Phase 6 validation repeat detail")
}
```

```{r phase-6-internal-validation-archived, eval=FALSE, include=FALSE}
# Archived full-data validation reference for a later high-memory rerun.
# Use this only on a machine that can handle the full pairwise distance load.
val_result_full <- clustMixType::validation_kproto(
  method = "silhouette",
  data = df_scenario,
  k = candidate_k,
  nstart = 5,
  type = "gower",
  kp_obj = "all",
  verbose = FALSE
)
```

```{r phase-6-stability}
stability_input <- copy(df_scenario)
stability_rows <- nrow(df_scenario)
stability_models <- kp_results[as.character(top_two_k)]
stability_repeat_results <- NULL
stability_bootstrap <- data.table()

if (identical(phase6_stability_mode, "consensus_sampled")) {
  stability_rows <- min(phase6_stability_sample_n, nrow(df_scenario))

  consensus_stability_outputs <- lapply(seq_len(phase6_stability_repeats), function(repeat_id) {
      set.seed(420 + repeat_id)
      stability_index <- sort(sample.int(nrow(df_scenario), stability_rows))
      stability_input_repeat <- copy(df_scenario[stability_index])

      repeat_models <- lapply(top_two_k, function(k_value) {
        set.seed(420 + repeat_id)
        clustMixType::kproto(
          x = stability_input_repeat,
          k = k_value,
          lambda = NULL,
          nstart = 5,
          type = "gower",
          keep.data = TRUE,
          verbose = FALSE
        )
      })
      names(repeat_models) <- as.character(top_two_k)

      repeat_stability_outputs <- lapply(top_two_k, function(k_value) {
        stab <- clustMixType::stability_kproto(
          object = repeat_models[[as.character(k_value)]],
          method = c("rand", "jaccard"),
          B = phase6_stability_bootstraps,
          verbose = FALSE,
          nstart = 5
        )

        bootstrap_dt <- as.data.table(stab[[2]])
        bootstrap_dt[, `:=`(k = k_value, repeat_id = repeat_id)]

        list(
          summary_dt = data.table(
            repeat_id = repeat_id,
            k = k_value,
            rand = as.numeric(stab[[1]]["R"]),
            jaccard = as.numeric(stab[[1]]["J"])
          ),
          bootstrap_dt = bootstrap_dt
        )
      })

      list(
        summary_dt = rbindlist(lapply(repeat_stability_outputs, `[[`, "summary_dt"), fill = TRUE),
        bootstrap_dt = rbindlist(lapply(repeat_stability_outputs, `[[`, "bootstrap_dt"), fill = TRUE)
      )
    })

  stability_repeat_results <- rbindlist(
    lapply(consensus_stability_outputs, `[[`, "summary_dt"),
    fill = TRUE
  )
  stability_bootstrap <- rbindlist(
    lapply(consensus_stability_outputs, `[[`, "bootstrap_dt"),
    fill = TRUE
  )

  stability_summary <- stability_repeat_results[, .(
    rand = mean(rand),
    jaccard = mean(jaccard),
    sd_rand = stats::sd(rand),
    sd_jaccard = stats::sd(jaccard),
    stability_repeats = .N,
    stable_repeat_share = round(mean(jaccard >= 0.6), 4),
    strong_repeat_share = round(mean(jaccard >= 0.75), 4),
    stable_jaccard = mean(jaccard >= 0.6) >= 0.5,
    strong_jaccard = mean(jaccard >= 0.75) >= 0.5
  ), by = k][order(-jaccard, -rand, k)]
} else {
  if (identical(phase6_stability_mode, "sampled")) {
    set.seed(42)
    stability_rows <- min(phase6_stability_sample_n, nrow(df_scenario))
    stability_index <- sort(sample.int(nrow(df_scenario), stability_rows))
    stability_input <- copy(df_scenario[stability_index])

    stability_models <- lapply(top_two_k, function(k_value) {
      set.seed(42)
      clustMixType::kproto(
        x = stability_input,
        k = k_value,
        lambda = NULL,
        nstart = 5,
        type = "gower",
        keep.data = TRUE,
        verbose = FALSE
      )
    })
    names(stability_models) <- as.character(top_two_k)
  }

  stability_results <- lapply(top_two_k, function(k_value) {
    set.seed(42)
    stab <- clustMixType::stability_kproto(
      object = stability_models[[as.character(k_value)]],
      method = c("rand", "jaccard"),
      B = phase6_stability_bootstraps,
      verbose = FALSE,
      nstart = 5
    )

    list(
      k = k_value,
      summary = stab[[1]],
      bootstrap = as.data.table(stab[[2]])
    )
  })

  stability_summary <- rbindlist(
    lapply(stability_results, function(result) {
      data.table(
        k = result$k,
        rand = as.numeric(result$summary["R"]),
        jaccard = as.numeric(result$summary["J"]),
        sd_rand = NA_real_,
        sd_jaccard = NA_real_,
        stability_repeats = 1L,
        stable_repeat_share = as.numeric(result$summary["J"]) >= 0.6,
        strong_repeat_share = as.numeric(result$summary["J"]) >= 0.75,
        stable_jaccard = as.numeric(result$summary["J"]) >= 0.6,
        strong_jaccard = as.numeric(result$summary["J"]) >= 0.75
      )
    })
  )[order(-jaccard, -rand, k)]

  stability_bootstrap <- rbindlist(
    lapply(stability_results, function(result) {
      bootstrap_dt <- copy(result$bootstrap)
      bootstrap_dt[, `:=`(k = result$k, repeat_id = 1L)]
      bootstrap_dt
    }),
    fill = TRUE
  )
}

stability_metadata <- data.table(
  setting = c("stability_mode", "stability_rows", "full_rows", "sample_fraction", "bootstrap_reps", "stability_repeats"),
  value = c(
    phase6_stability_mode,
    stability_rows,
    nrow(df_scenario),
    round(stability_rows / nrow(df_scenario), 4),
    phase6_stability_bootstraps,
    phase6_stability_repeats
  )
)

fwrite(stability_summary, file.path(output_dir, "phase6_stability_summary.csv"))
fwrite(stability_bootstrap, file.path(output_dir, "phase6_stability_bootstrap.csv"))
fwrite(stability_metadata, file.path(output_dir, "phase6_stability_metadata.csv"))
if (!is.null(stability_repeat_results)) {
  fwrite(stability_repeat_results, file.path(output_dir, "phase6_stability_repeats.csv"))
}

knitr::kable(stability_summary, caption = "Phase 6 stability summary")
knitr::kable(stability_metadata, caption = "Phase 6 stability run settings")
if (!is.null(stability_repeat_results)) {
  knitr::kable(stability_repeat_results, caption = "Phase 6 stability repeat detail")
}
```

```{r phase-6-jaccard-ci}
# v2.0: Compute Jaccard mean, SD, and range across bootstrap repeats.
# Under consensus_16gb (3 repeats) this gives a variance estimate.
# Under local_safe (1 repeat) this falls back gracefully.
if (!is.null(stability_repeat_results) &&
    nrow(stability_repeat_results) > 0 &&
    "jaccard" %in% names(stability_repeat_results)) {

  jaccard_ci <- stability_repeat_results[, .(
    jaccard_mean = round(mean(jaccard, na.rm = TRUE), 4),
    jaccard_sd   = round(sd(jaccard,   na.rm = TRUE), 4),
    jaccard_min  = round(min(jaccard,  na.rm = TRUE), 4),
    jaccard_max  = round(max(jaccard,  na.rm = TRUE), 4),
    n_repeats    = .N
  ), by = k]

  fwrite(jaccard_ci, file.path(output_dir, "phase6_jaccard_ci.csv"))
  knitr::kable(
    jaccard_ci,
    caption = "Phase 6 Jaccard stability with variance across bootstrap repeats"
  )
} else {
  cat("Single bootstrap repeat --- Jaccard CI not available under local_safe.\n")
}
```

```{r phase-6-stability-archived, eval=FALSE, include=FALSE}
# Archived full-data stability reference for a later high-memory rerun.
stability_results_full <- lapply(top_two_k, function(k_value) {
  set.seed(42)
  clustMixType::stability_kproto(
    object = kp_results[[as.character(k_value)]],
    method = c("rand", "jaccard"),
    B = 50,
    verbose = FALSE,
    nstart = 5
  )
})
```

```{r phase-6-select-k}
# v2.0: Verbose k-selection with explicit diagnostics.
# The original silent fallback could select a k without any trace of
# which conditions fired or whether NA extraction caused a bypass.

validation_optimal_k_raw <- validation_table$k[1]
silhouette_at_optimal    <- validation_table$silhouette[1]

cat("=== Phase 6 k-Selection Diagnostics ===\n")
cat("Silhouette-optimal k        :", validation_optimal_k_raw, "\n")
cat("Silhouette at optimal k     :", round(silhouette_at_optimal, 4), "\n")
cat("Silhouette threshold (0.30) :",
    ifelse(silhouette_at_optimal >= 0.30, "MET", "NOT MET"), "\n\n")
cat("Stability available for k   :",
    paste(stability_summary$k, collapse = ", "), "\n")

selected_stability <- stability_summary[
  k == validation_optimal_k_raw, jaccard
]

cat("Jaccard at silhouette-optimal k :",
    ifelse(length(selected_stability) == 1 && !is.na(selected_stability),
           round(selected_stability, 4), "NOT FOUND / NA"), "\n")
cat("Jaccard threshold (0.40)        :",
    ifelse(length(selected_stability) == 1 &&
             !is.na(selected_stability) &&
             selected_stability >= 0.40, "MET", "NOT MET"), "\n\n")

fallback_triggered <- FALSE
if (length(selected_stability) == 1 &&
    !is.na(selected_stability) &&
    selected_stability < 0.5 &&
    nrow(stability_summary) > 1) {

  alt_k       <- stability_summary[order(-jaccard, -rand, k)][1, k]
  alt_jaccard <- stability_summary[order(-jaccard, -rand, k)][1, jaccard]

  cat("Fallback condition met. Stability-optimal k:", alt_k,
      "| Jaccard:", round(alt_jaccard, 4), "\n")

  if (!is.na(alt_jaccard) && alt_jaccard > selected_stability) {
    selected_k         <- alt_k
    fallback_triggered <- TRUE
    cat("Fallback ACTIVATED: selected_k changed to", selected_k, "\n\n")
  } else {
    selected_k <- validation_optimal_k_raw
    cat("Fallback NOT activated: no better Jaccard available.",
        "Retaining k =", selected_k, "\n\n")
  }
} else {
  selected_k <- validation_optimal_k_raw
  cat("Fallback condition not met. Using silhouette-optimal k =",
      selected_k, "\n\n")
}
cat("FINAL selected_k :", selected_k, "\n")
cat("========================================\n")

selected_model <- kp_results[[as.character(selected_k)]]
selected_jaccard <- if (selected_k %in% stability_summary$k) {
  stability_summary[k == selected_k, round(jaccard, 4)]
} else {
  NA_real_
}

# v2.0: Explicit quality gate — fires a warning and builds a status table
# whenever silhouette or Jaccard fall below published thresholds.
selected_silhouette <- validation_table[k == selected_k, silhouette]
selected_jaccard_v2 <- if (selected_k %in% stability_summary$k) {
  stability_summary[k == selected_k, jaccard]
} else {
  NA_real_
}

sil_pass     <- !is.na(selected_silhouette) && selected_silhouette >= 0.30
jaccard_pass <- !is.na(selected_jaccard_v2) && selected_jaccard_v2 >= 0.40

quality_gate <- data.table(
  metric = c("Silhouette width", "Jaccard stability"),
  observed = c(
    round(selected_silhouette, 4),
    round(selected_jaccard_v2, 4)
  ),
  threshold = c(0.30, 0.40),
  pass = c(sil_pass, jaccard_pass),
  interpretation = c(
    ifelse(sil_pass,
           "Acceptable cluster separation",
           "Weak separation — exploratory only"),
    ifelse(jaccard_pass,
           "Acceptable bootstrap stability",
           "Poor stability — exploratory only")
  )
)

fwrite(quality_gate, file.path(output_dir, "phase6_quality_gate.csv"))
knitr::kable(quality_gate, caption = "Phase 6 cluster quality gate (v2.0)")

if (!sil_pass || !jaccard_pass) {
  warning(paste0(
    "\n\n*** CLUSTER QUALITY WARNING (v2.0) ***\n",
    "  Silhouette = ", round(selected_silhouette, 4),
    " (threshold >= 0.30): ", ifelse(sil_pass, "PASS", "FAIL"), "\n",
    "  Jaccard    = ", round(selected_jaccard_v2, 4),
    " (threshold >= 0.40): ", ifelse(jaccard_pass, "PASS", "FAIL"), "\n",
    "  selected_k = ", selected_k,
    " solution is EXPLORATORY.\n",
    "  Do not present clusters as stable accident archetypes.\n",
    "***********************************************\n"
  ))
}

selection_summary <- data.table(
  metric = c(
    "Validation method",
    "Validation mode",
    "Validation rows",
    "Stability mode",
    "Stability rows",
    "Stability bootstraps",
    "Validation-optimal k",
    "Top two k values for stability",
    "Selected k",
    "Selected k Jaccard"
  ),
  value = c(
    "silhouette",
    phase6_validation_mode,
    validation_rows,
    phase6_stability_mode,
    stability_rows,
    phase6_stability_bootstraps,
    validation_optimal_k,
    paste(top_two_k, collapse = ", "),
    selected_k,
    selected_jaccard
  )
)

fwrite(selection_summary, file.path(output_dir, "phase6_selection_summary.csv"))

knitr::kable(selection_summary, caption = "Phase 6 k-selection summary")
```

```{r phase-6-reduced-validation}
# v2.0: Validate reduced-feature k-prototypes candidates.
# Compares silhouette scores with and without Pressure(in)/Visibility(mi).
# v2.0: Mirror the same sampling strategy used in the main Phase 6 validation
# to stay within memory limits. Uses phase6_validation_sample_n rows.
set.seed(42)
reduced_val_rows <- min(phase6_validation_sample_n, nrow(df_scenario_reduced))
reduced_val_input <- if (identical(phase6_validation_mode, "full")) {
  copy(df_scenario_reduced)
} else {
  reduced_sampled_idx <- sort(sample.int(nrow(df_scenario_reduced), reduced_val_rows))
  copy(df_scenario_reduced[reduced_sampled_idx])
}

cat("Reduced-feature validation sample size:", nrow(reduced_val_input), "\n")

validation_reduced_raw <- clustMixType::validation_kproto(
  method  = "silhouette",
  data    = reduced_val_input,
  k       = candidate_k,
  nstart  = 5,
  type    = "gower",
  kp_obj  = "all",
  verbose = FALSE
)

validation_reduced_table <- data.table(
  k          = as.integer(names(validation_reduced_raw$indices)),
  silhouette = as.numeric(validation_reduced_raw$indices)
)
validation_reduced_table[, feature_set := "reduced"]

validation_full_labeled <- copy(
  validation_table[, .(k, silhouette)]
)[, feature_set := "full"]

feature_silhouette_comparison <- rbind(
  validation_full_labeled,
  validation_reduced_table[, .(k, silhouette, feature_set)]
)
setorder(feature_silhouette_comparison, k, feature_set)

fwrite(
  feature_silhouette_comparison,
  file.path(output_dir, "phase6_feature_silhouette_comparison.csv")
)

knitr::kable(
  feature_silhouette_comparison,
  caption = "Phase 6 silhouette: full vs reduced feature set"
)
```

```{r phase-6-cluster-profiles}
profile_plot_path <- file.path(output_dir, "phase6_cluster_profiles.pdf")
grDevices::pdf(profile_plot_path, width = 12, height = 8)
clustMixType::clprofiles(selected_model, df_scenario)
grDevices::dev.off()

selected_cluster_data <- copy(df_scenario)
selected_cluster_data[, cluster := factor(selected_model$cluster)]

cluster_profile_summary <- selected_cluster_data[, .(
  dominant_hour_band = mode_value(hour_band),
  dominant_weather_group = mode_value(weather_group),
  dominant_weekend = mode_value(weekend),
  dominant_sunrise_sunset = mode_value(Sunrise_Sunset),
  dominant_junction = mode_value(Junction),
  dominant_traffic_signal = mode_value(Traffic_Signal),
  mean_temperature = round(mean(`Temperature(F)`), 2),
  mean_visibility = round(mean(`Visibility(mi)`), 2),
  mean_log_duration = round(mean(log_Duration), 3)
), by = cluster]

cluster_profile_summary[, working_name := dplyr::case_when(
  dominant_weather_group == "Snow_Ice" ~ "Winter Adverse",
  dominant_sunrise_sunset == "Night" & dominant_weather_group %in% c("Fog_Haze", "Rain", "Thunderstorm") ~ "Night Low-Visibility",
  dominant_weekend == "TRUE" & dominant_traffic_signal == "TRUE" ~ "Weekend Surface Road",
  dominant_hour_band %in% c("AM_Rush", "PM_Rush") & dominant_junction == "TRUE" ~ "Rush Corridor",
  TRUE ~ paste("Scenario Cluster", cluster)
)]

fwrite(cluster_profile_summary, file.path(output_dir, "phase6_cluster_profile_summary.csv"))

knitr::kable(cluster_profile_summary, caption = "Phase 6 cluster profile summary")
```

```{r phase-6-summary}
phase_6_summary <- data.table(
  checkpoint = c(
    "Validation indices saved",
    "Validation metadata saved",
    "Stability summary saved",
    "Stability metadata saved",
    "Selected k assigned",
    "Cluster profile PDF saved",
    "Working names assigned"
  ),
  status = c(
    file.exists(file.path(output_dir, "phase6_validation_indices.csv")),
    file.exists(file.path(output_dir, "phase6_validation_metadata.csv")),
    file.exists(file.path(output_dir, "phase6_stability_summary.csv")),
    file.exists(file.path(output_dir, "phase6_stability_metadata.csv")),
    !is.null(selected_model),
    file.exists(file.path(output_dir, "phase6_cluster_profiles.pdf")),
    all(nzchar(cluster_profile_summary$working_name))
  )
)

fwrite(phase_6_summary, file.path(output_dir, "phase6_summary.csv"))

knitr::kable(phase_6_summary, caption = "Phase 6 readiness summary")
```

# Phase 7: DBSCAN Spatial Hotspot Clustering

This phase clusters the projected Long Island coordinates with DBSCAN so the workflow can identify corridor-style hotspot zones.

```{r phase-7-parameter-selection}
knn_k <- 5L
dbscan_minpts_grid <- if (run_profile == "full_srs") c(15L, 25L, 40L, 50L, 75L, 100L) else c(40L, 50L, 75L, 100L)
dbscan_eps_grid <- seq(0.3, 2.0, by = 0.1)

knn_dist <- dbscan::kNNdist(df_spatial, k = knn_k)
knn_sorted <- sort(knn_dist)
phase7_total_rows <- nrow(df_spatial)
phase7_target_cluster_n <- 8L
phase7_target_noise_pct <- 17.5
phase7_max_cluster_share_cap <- 60

knn_profile <- data.table(
  quantile = c("50%", "75%", "90%", "95%", "97.5%", "99%"),
  value_km = round(as.numeric(quantile(knn_dist, probs = c(0.5, 0.75, 0.9, 0.95, 0.975, 0.99))), 4)
)

eps_trials <- rbindlist(
  lapply(dbscan_minpts_grid, function(minPts_value) {
    rbindlist(lapply(dbscan_eps_grid, function(eps_value) {
      db_trial <- dbscan::dbscan(df_spatial, eps = eps_value, minPts = minPts_value)
      cluster_counts <- table(db_trial$cluster)
      non_noise_sizes <- cluster_counts[names(cluster_counts) != "0"]
      non_noise_clusters <- sum(as.integer(names(cluster_counts)) > 0)
      noise_pct <- round(100 * mean(db_trial$cluster == 0), 2)

      data.table(
        eps = round(eps_value, 2),
        minPts = minPts_value,
        non_noise_clusters = non_noise_clusters,
        noise_pct = noise_pct,
        max_cluster_size = if (length(non_noise_sizes)) max(as.integer(non_noise_sizes)) else 0L,
        max_cluster_share_pct = if (length(non_noise_sizes)) round(100 * max(as.integer(non_noise_sizes)) / phase7_total_rows, 2) else 0,
        size_balance_ok = if (length(non_noise_sizes)) round(100 * max(as.integer(non_noise_sizes)) / phase7_total_rows, 2) <= phase7_max_cluster_share_cap else FALSE,
        acceptable = non_noise_clusters >= 3 & non_noise_clusters <= 15 & noise_pct >= 5 & noise_pct <= 30 &
          if (length(non_noise_sizes)) round(100 * max(as.integer(non_noise_sizes)) / phase7_total_rows, 2) <= phase7_max_cluster_share_cap else FALSE
      )
    }))
  }),
  fill = TRUE
)

acceptable_trials <- eps_trials[acceptable == TRUE]

if (nrow(acceptable_trials) > 0) {
  selected_trial <- acceptable_trials[
    order(abs(non_noise_clusters - phase7_target_cluster_n), abs(noise_pct - phase7_target_noise_pct), max_cluster_share_pct, -minPts, eps)
  ][1]
  selection_reason <- "Accepted trial balancing cluster count, noise level, and hotspot size concentration"
} else {
  selected_trial <- eps_trials[
    order(abs(non_noise_clusters - phase7_target_cluster_n), abs(noise_pct - phase7_target_noise_pct), max_cluster_share_pct, -minPts, eps)
  ][1]
  selection_reason <- "Fallback trial closest to balanced cluster count, noise level, and hotspot size concentration"
}

eps_chosen <- selected_trial$eps[[1]]
minPts_chosen <- selected_trial$minPts[[1]]

parameter_summary <- data.table(
  metric = c(
    "kNN distance k",
    "DBSCAN minPts grid",
    "DBSCAN eps grid",
    "Largest cluster share cap (%)",
    "DBSCAN minPts",
    "Chosen eps (km)",
    "Chosen non-noise clusters",
    "Chosen noise pct",
    "Chosen largest cluster share pct",
    "Selection reason"
  ),
  value = c(
    knn_k,
    paste(dbscan_minpts_grid, collapse = ", "),
    paste(range(dbscan_eps_grid), collapse = " to "),
    phase7_max_cluster_share_cap,
    minPts_chosen,
    eps_chosen,
    selected_trial$non_noise_clusters[[1]],
    selected_trial$noise_pct[[1]],
    selected_trial$max_cluster_share_pct[[1]],
    selection_reason
  )
)

fwrite(knn_profile, file.path(output_dir, "phase7_knn_profile.csv"))
fwrite(eps_trials, file.path(output_dir, "phase7_eps_trials.csv"))
fwrite(parameter_summary, file.path(output_dir, "phase7_parameter_summary.csv"))

knitr::kable(knn_profile, caption = "Phase 7 kNN distance profile")
knitr::kable(eps_trials, caption = "Phase 7 DBSCAN eps trials")
knitr::kable(parameter_summary, caption = "Phase 7 selected DBSCAN settings")
```

```{r phase-7-knn-plot, fig.height=4.5, fig.width=7}
knn_plot_data <- data.table(
  point_order = seq_along(knn_sorted),
  distance_km = knn_sorted
)

knn_plot <- ggplot2::ggplot(
  knn_plot_data,
  ggplot2::aes(x = point_order, y = distance_km)
) +
  ggplot2::geom_line(color = "#1f78b4", linewidth = 0.6) +
  ggplot2::geom_hline(yintercept = eps_chosen, linetype = "dashed", color = "#e31a1c") +
  ggplot2::labs(
    title = "Phase 7 kNN Distance Plot",
    x = "Points sorted by 5-NN distance",
    y = "5-NN distance (km)"
  ) +
  ggplot2::theme_minimal()

print(knn_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase7_knn_distance_plot.png"),
  plot = knn_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

```{r phase-7-dbscan-run}
db_result <- dbscan::dbscan(df_spatial, eps = eps_chosen, minPts = minPts_chosen)

df_spatial_clusters <- copy(df_spatial)
df_spatial_clusters[, spatial_cluster := as.integer(db_result$cluster)]

spatial_cluster_sizes <- df_spatial_clusters[, .N, by = spatial_cluster][order(spatial_cluster)]
largest_non_noise_size <- max(spatial_cluster_sizes[spatial_cluster > 0, N], na.rm = TRUE)
largest_non_noise_share_pct <- round(100 * largest_non_noise_size / nrow(df_spatial_clusters), 2)
solution_summary <- data.table(
  metric = c(
    "Rows clustered",
    "Noise pct",
    "Non-noise clusters",
    "Largest cluster size",
    "Largest cluster share pct",
    "Meets SRS cluster-count target",
    "Meets SRS noise target",
    "Meets hotspot-balance target"
  ),
  value = c(
    nrow(df_spatial_clusters),
    round(100 * mean(df_spatial_clusters$spatial_cluster == 0), 2),
    uniqueN(df_spatial_clusters$spatial_cluster[df_spatial_clusters$spatial_cluster > 0]),
    largest_non_noise_size,
    largest_non_noise_share_pct,
    uniqueN(df_spatial_clusters$spatial_cluster[df_spatial_clusters$spatial_cluster > 0]) >= 3 &
      uniqueN(df_spatial_clusters$spatial_cluster[df_spatial_clusters$spatial_cluster > 0]) <= 15,
    round(100 * mean(df_spatial_clusters$spatial_cluster == 0), 2) >= 5 &
      round(100 * mean(df_spatial_clusters$spatial_cluster == 0), 2) <= 30,
    largest_non_noise_share_pct <= phase7_max_cluster_share_cap
  )
)

if (!any(df_spatial_clusters$spatial_cluster > 0)) {
  stop("Phase 7 DBSCAN produced only noise points; adjust eps or minPts.")
}

fwrite(spatial_cluster_sizes, file.path(output_dir, "phase7_dbscan_cluster_sizes.csv"))
fwrite(solution_summary, file.path(output_dir, "phase7_solution_summary.csv"))

knitr::kable(spatial_cluster_sizes, caption = "Phase 7 spatial cluster sizes")
knitr::kable(solution_summary, caption = "Phase 7 DBSCAN solution summary")
```

```{r phase-7-cluster-centroids}
# v2.0: Compute centroid lat/lng for each non-noise spatial cluster.
# Export so corridor assignments can be manually verified against a map.
# Reference point: lat0 = 40.8 N, lng0 = -73.2 W (same as projection).
cluster_centroids <- df_spatial_clusters[spatial_cluster > 0, .(
  centroid_x_km = round(mean(x_km), 4),
  centroid_y_km = round(mean(y_km), 4),
  centroid_lat  = round(
    40.8 + mean(y_km) / 110.574, 5
  ),
  centroid_lng  = round(
    -73.2 + mean(x_km) / (111.32 * cos(40.8 * pi / 180)), 5
  ),
  n_accidents = .N
), by = spatial_cluster]

setorder(cluster_centroids, spatial_cluster)

fwrite(
  cluster_centroids,
  file.path(output_dir, "phase7_cluster_centroids.csv")
)

knitr::kable(
  cluster_centroids,
  caption = paste(
    "Phase 7 spatial cluster centroids (v2.0).",
    "Verify each centroid against the named corridor.",
    "lat0 = 40.8 N, lng0 = 73.2 W."
  )
)
```

```{r phase-7-eps-sensitivity}
# v2.0: Test eps sensitivity at selected +/- 0.1 km.
# Confirms the 12-cluster solution is not highly sensitive to eps.
eps_test_values <- c(
  eps_chosen - 0.1,
  eps_chosen,
  eps_chosen + 0.1
)

eps_sensitivity_table <- rbindlist(lapply(eps_test_values, function(eps_test) {
  res <- dbscan::dbscan(
    df_spatial[, .(x_km, y_km)],
    eps    = eps_test,
    minPts = minPts_chosen
  )
  n_clusters <- length(unique(res$cluster[res$cluster > 0]))
  noise_pct  <- round(100 * mean(res$cluster == 0), 2)
  max_share  <- if (sum(res$cluster > 0) > 0) {
    round(100 * max(table(res$cluster[res$cluster > 0])) /
            sum(res$cluster > 0), 2)
  } else NA_real_

  data.table(
    eps                   = eps_test,
    n_clusters            = n_clusters,
    noise_pct             = noise_pct,
    max_cluster_share_pct = max_share,
    within_srs_targets    = (
      n_clusters >= 3 & n_clusters <= 15 &
      noise_pct  >= 5 & noise_pct  <= 30 &
      max_share  <= 60
    )
  )
}))

fwrite(
  eps_sensitivity_table,
  file.path(output_dir, "phase7_eps_sensitivity.csv")
)

knitr::kable(
  eps_sensitivity_table,
  caption = "Phase 7 DBSCAN eps sensitivity (+/- 0.1 km from selected eps)"
)
```

```{r phase-7-corridor-assignment}
lat0 <- 40.8
lng0 <- -73.2

non_noise_assignments <- df_spatial_clusters[spatial_cluster > 0, .(
  cluster_size = .N,
  centroid_x_km = mean(x_km),
  centroid_y_km = mean(y_km),
  x_span_km = diff(range(x_km)),
  y_span_km = diff(range(y_km))
), by = spatial_cluster][
  , `:=`(
    centroid_lat = round((centroid_y_km / 110.574) + lat0, 5),
    centroid_lng = round((centroid_x_km / (111.32 * cos(lat0 * pi / 180))) + lng0, 5),
    shape_hint = fifelse(x_span_km >= y_span_km, "east_west", "north_south")
  )
]

non_noise_assignments[, corridor_label := dplyr::case_when(
  centroid_lat >= 40.74 & centroid_lat <= 40.76 ~ "Long Island Expressway (I-495)",
  centroid_lat >= 40.68 & centroid_lat <= 40.72 ~ "Southern State Parkway",
  centroid_lat >= 40.78 & centroid_lat <= 40.82 ~ "Northern State Parkway",
  centroid_lat >= 40.64 & centroid_lat <= 40.68 ~ "Sunrise Highway",
  centroid_lat >= 40.72 & centroid_lat <= 40.76 & shape_hint == "north_south" ~ "Meadowbrook / Wantagh Pkwy",
  TRUE ~ "Surface Road Cluster"
)]

fwrite(non_noise_assignments, file.path(output_dir, "phase7_corridor_assignments.csv"))

knitr::kable(non_noise_assignments, caption = "Phase 7 corridor assignment summary")
```

```{r phase-7-spatial-map, fig.height=5.5, fig.width=8}
df_spatial_plot <- copy(df_spatial_clusters)
df_spatial_plot[, cluster_label := factor(spatial_cluster)]

cluster_levels <- sort(unique(df_spatial_plot$cluster_label))
non_noise_levels <- setdiff(as.character(cluster_levels), "0")
cluster_colors <- c(
  "0" = "#bdbdbd",
  setNames(grDevices::hcl.colors(length(non_noise_levels), palette = "Dark 3"), non_noise_levels)
)

spatial_map <- ggplot2::ggplot(
  df_spatial_plot,
  ggplot2::aes(x = x_km, y = y_km, color = cluster_label)
) +
  ggplot2::geom_point(size = 0.5, alpha = 0.6) +
  ggplot2::geom_text(
    data = non_noise_assignments,
    ggplot2::aes(x = centroid_x_km, y = centroid_y_km, label = spatial_cluster),
    inherit.aes = FALSE,
    size = 3.2,
    fontface = "bold",
    color = "black"
  ) +
  ggplot2::scale_color_manual(values = cluster_colors) +
  ggplot2::labs(
    title = "Phase 7 DBSCAN Spatial Hotspots",
    x = "Projected x (km)",
    y = "Projected y (km)",
    color = "Cluster"
  ) +
  ggplot2::theme_minimal()

print(spatial_map)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase7_spatial_hotspot_map.png"),
  plot = spatial_map,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

```{r phase-7-spatial-density-heatmap, fig.height=5.5, fig.width=8}
spatial_density_heatmap <- ggplot2::ggplot(
  df_spatial_plot,
  ggplot2::aes(
    x = x_km,
    y = y_km,
    fill = after_stat(log10(count + 1))
  )
) +
  ggplot2::geom_bin2d(bins = 100) +
  ggplot2::geom_point(
    data = non_noise_assignments,
    ggplot2::aes(x = centroid_x_km, y = centroid_y_km),
    inherit.aes = FALSE,
    shape = 21,
    size = 2.6,
    stroke = 0.5,
    fill = "white",
    color = "black"
  ) +
  ggplot2::geom_text(
    data = non_noise_assignments,
    ggplot2::aes(x = centroid_x_km, y = centroid_y_km, label = spatial_cluster),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold",
    nudge_y = 0.6
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "C",
    name = "log10(bin\ncount + 1)"
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Phase 7 Long Island Spatial Density Heatmap",
    subtitle = "Heat intensity shows accident concentration; numbered markers show hotspot centroids",
    x = "Projected x (km)",
    y = "Projected y (km)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

print(spatial_density_heatmap)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase7_spatial_density_heatmap.png"),
  plot = spatial_density_heatmap,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

```{r phase-7-summary}
phase_7_summary <- data.table(
  checkpoint = c(
    "kNN distance plot saved",
    "eps trials saved",
    "DBSCAN solution created",
    "Corridor assignments saved",
    "Spatial hotspot map saved",
    "Spatial density heatmap saved"
  ),
  status = c(
    file.exists(file.path(output_dir, "phase7_knn_distance_plot.png")),
    file.exists(file.path(output_dir, "phase7_eps_trials.csv")),
    exists("db_result"),
    file.exists(file.path(output_dir, "phase7_corridor_assignments.csv")),
    file.exists(file.path(output_dir, "phase7_spatial_hotspot_map.png")),
    file.exists(file.path(output_dir, "phase7_spatial_density_heatmap.png"))
  )
)

fwrite(phase_7_summary, file.path(output_dir, "phase7_summary.csv"))

knitr::kable(phase_7_summary, caption = "Phase 7 readiness summary")
```

# Phase 8: Post-Clustering Interpretation

This phase joins the scenario and spatial cluster labels back to the working data and builds the interpretation tables used for the final report.

```{r phase-8-label-joining}
kp_final <- selected_model
db_final <- db_result

if (length(kp_final$cluster) != nrow(df_phase3)) {
  stop("Scenario cluster labels do not match the Phase 3 data rows.")
}

if (length(db_final$cluster) != nrow(df_phase3)) {
  stop("Spatial cluster labels do not match the Phase 3 data rows.")
}

df_clustered <- copy(df_phase3)
df_clustered[, scenario_cluster := as.integer(kp_final$cluster)]
df_clustered[, spatial_cluster := as.integer(db_final$cluster)]

label_join_check <- data.table(
  metric = c(
    "Rows in clustered data",
    "Scenario labels",
    "Spatial labels",
    "Scenario label NAs",
    "Spatial label NAs"
  ),
  value = c(
    nrow(df_clustered),
    length(kp_final$cluster),
    length(db_final$cluster),
    sum(is.na(df_clustered$scenario_cluster)),
    sum(is.na(df_clustered$spatial_cluster))
  )
)

if (any(label_join_check[metric %in% c("Scenario label NAs", "Spatial label NAs"), value > 0])) {
  stop("Phase 8 label joining produced missing cluster labels.")
}

fwrite(label_join_check, file.path(output_dir, "phase8_label_join_check.csv"))

knitr::kable(label_join_check, caption = "Phase 8 label join check")
```

```{r phase-8-severity-profile}
dataset_sev4_pct <- round(100 * mean(as.integer(df_clustered$Severity) == 4), 2)
dataset_sev34_pct <- round(100 * mean(as.integer(df_clustered$Severity) %in% c(3, 4)), 2)

scenario_severity_profile <- df_clustered[, .(
  n = .N,
  sev1_pct = round(100 * mean(as.integer(Severity) == 1), 2),
  sev2_pct = round(100 * mean(as.integer(Severity) == 2), 2),
  sev3_pct = round(100 * mean(as.integer(Severity) == 3), 2),
  sev4_pct = round(100 * mean(as.integer(Severity) == 4), 2),
  sev34_pct = round(100 * mean(as.integer(Severity) %in% c(3, 4)), 2)
), by = scenario_cluster][order(scenario_cluster)]

scenario_severity_profile[, high_risk_archetype := (
  sev4_pct >= (1.5 * dataset_sev4_pct) |
    sev34_pct == max(sev34_pct)
)]

# v2.0: Add explicit Sev4 rate and ratio-to-baseline columns
scenario_severity_profile[, sev4_rate       := sev4_pct]
scenario_severity_profile[, sev4_vs_baseline := round(sev4_pct / dataset_sev4_pct, 2)]

# v2.0: Prepend a baseline row (scenario_cluster = 0) for easy comparison
baseline_severity_row <- data.table(
  scenario_cluster = 0L,
  n                = nrow(df_clustered),
  sev1_pct         = round(100 * mean(as.integer(df_clustered$Severity) == 1), 2),
  sev2_pct         = round(100 * mean(as.integer(df_clustered$Severity) == 2), 2),
  sev3_pct         = round(100 * mean(as.integer(df_clustered$Severity) == 3), 2),
  sev4_pct         = dataset_sev4_pct,
  sev34_pct        = dataset_sev34_pct,
  high_risk_archetype = FALSE,
  sev4_rate        = dataset_sev4_pct,
  sev4_vs_baseline = 1.00
)

severity_profile_v2 <- rbind(baseline_severity_row, scenario_severity_profile, fill = TRUE)

fwrite(severity_profile_v2, file.path(output_dir, "phase8_severity_profile_v2.csv"))
fwrite(scenario_severity_profile, file.path(output_dir, "phase8_scenario_severity_profile.csv"))

knitr::kable(
  severity_profile_v2,
  caption = paste(
    "Phase 8 severity profile (v2.0).",
    "Row 0 = dataset baseline.",
    "sev4_vs_baseline = cluster Sev4 rate / baseline Sev4 rate."
  )
)
```

```{r phase-8-county-profile}
scenario_county_profile <- df_clustered[, .(
  n = .N,
  Nassau_pct = round(100 * mean(County == "Nassau"), 2),
  Suffolk_pct = round(100 * mean(County == "Suffolk"), 2),
  Queens_pct = round(100 * mean(County == "Queens"), 2),
  dominant_county = mode_value(County)
), by = scenario_cluster][order(scenario_cluster)]

fwrite(scenario_county_profile, file.path(output_dir, "phase8_scenario_county_profile.csv"))

knitr::kable(scenario_county_profile, caption = "Phase 8 county composition by scenario cluster")
```

```{r phase-8-weather-time-profile}
scenario_weather_time_profile <- df_clustered[, .(
  n = .N,
  dominant_weather_group = mode_value(weather_group),
  dominant_weather_share = round(100 * dominant_share(weather_group), 2),
  dominant_hour_band = mode_value(hour_band),
  dominant_hour_share = round(100 * dominant_share(hour_band), 2),
  night_pct = round(100 * mean(Sunrise_Sunset == "Night"), 2),
  weekend_pct = round(100 * mean(weekend), 2),
  mean_temperature = round(mean(`Temperature(F)`), 2),
  mean_visibility = round(mean(`Visibility(mi)`), 2),
  mean_wind_speed = round(mean(`Wind_Speed(mph)`), 2),
  junction_pct = round(100 * mean(Junction == "TRUE"), 2),
  traffic_signal_pct = round(100 * mean(Traffic_Signal == "TRUE"), 2),
  crossing_pct = round(100 * mean(Crossing == "TRUE"), 2),
  stop_pct = round(100 * mean(Stop == "TRUE"), 2)
), by = scenario_cluster][order(scenario_cluster)]

fwrite(scenario_weather_time_profile, file.path(output_dir, "phase8_scenario_weather_time_profile.csv"))

knitr::kable(scenario_weather_time_profile, caption = "Phase 8 weather and time profile by scenario cluster")
```

```{r phase-8-spatial-scenario-crosstab}
scenario_name_lookup <- copy(cluster_profile_summary)[
  , .(
    scenario_cluster = as.integer(as.character(cluster)),
    working_name
  )
]

spatial_scenario_crosstab <- df_clustered[spatial_cluster > 0, .N, by = .(spatial_cluster, scenario_cluster)][
  order(spatial_cluster, -N, scenario_cluster)
][
  , spatial_cluster_pct := round(100 * N / sum(N), 2), by = spatial_cluster
]

spatial_scenario_crosstab <- merge(
  spatial_scenario_crosstab,
  non_noise_assignments[, .(spatial_cluster, corridor_label)],
  by = "spatial_cluster",
  all.x = TRUE
)

spatial_scenario_crosstab <- merge(
  spatial_scenario_crosstab,
  scenario_name_lookup,
  by = "scenario_cluster",
  all.x = TRUE
)

spatial_scenario_dominance <- spatial_scenario_crosstab[
  order(spatial_cluster, -N, scenario_cluster)
][
  , .SD[1], by = spatial_cluster
][
  , .(
    spatial_cluster,
    corridor_label,
    dominant_scenario_cluster = scenario_cluster,
    dominant_scenario_name = working_name,
    dominant_count = N,
    dominant_pct = spatial_cluster_pct
  )
]

fwrite(spatial_scenario_crosstab, file.path(output_dir, "phase8_spatial_scenario_crosstab.csv"))
fwrite(spatial_scenario_dominance, file.path(output_dir, "phase8_spatial_scenario_dominance.csv"))

knitr::kable(spatial_scenario_crosstab, caption = "Phase 8 spatial-scenario cross-tabulation")
knitr::kable(spatial_scenario_dominance, caption = "Phase 8 dominant scenario by spatial hotspot")
```

```{r phase-8-final-naming}
scenario_final_names <- merge(
  scenario_weather_time_profile,
  scenario_severity_profile[, .(scenario_cluster, sev4_pct, high_risk_archetype)],
  by = "scenario_cluster",
  all.x = TRUE
)

scenario_final_names <- merge(
  scenario_final_names,
  scenario_name_lookup,
  by = "scenario_cluster",
  all.x = TRUE
)

# v2.0: Diagnostic print — inspect all profile values before naming
# so any cluster falling through to the generic fallback is visible.
cat("=== Phase 8 Cluster Profile Values for Naming Diagnostics ===\n")
print(scenario_final_names[, .(
  scenario_cluster, dominant_hour_band, dominant_weather_group,
  night_pct, weekend_pct, junction_pct, traffic_signal_pct
)])
cat("=============================================================\n")

scenario_final_names[, final_name := dplyr::case_when(
  # Tier 1: Extreme weather
  dominant_weather_group == "Snow_Ice" ~
    "Winter Adverse --- Snow and Ice",

  # Tier 2: Night + hazardous visibility
  night_pct >= 50 &
    dominant_weather_group %in% c("Fog_Haze", "Rain", "Thunderstorm") ~
    "Night Low-Visibility --- Fog and Rain",

  # Tier 3: Night / early morning (any weather)
  night_pct >= 50 & dominant_hour_band == "Early_AM" ~
    paste("Night / Early AM ---", gsub("_", " ", dominant_weather_group)),

  # Tier 4: Weekend surface road
  weekend_pct >= 40 & traffic_signal_pct >= 8 ~
    "Weekend Surface Road --- Signal Heavy",

  # Tier 5: Rush-hour corridor
  dominant_hour_band %in% c("AM_Rush", "PM_Rush") & junction_pct >= 10 ~
    paste(
      gsub("_", " ", dominant_weather_group),
      gsub("_", " ", dominant_hour_band),
      "--- Junction Heavy"
    ),

  # Tier 6 (v2.0): Cloudy off-peak daytime — captures generic midday clusters
  dominant_weather_group == "Cloudy" &
    dominant_hour_band %in% c("Midday", "Evening") ~
    paste("Cloudy Off-Peak ---", gsub("_", " ", dominant_hour_band)),

  # Tier 7 (v2.0): Clear off-peak daytime
  dominant_weather_group == "Clear" &
    dominant_hour_band %in% c("Midday", "Evening") ~
    paste("Clear Off-Peak ---", gsub("_", " ", dominant_hour_band)),

  # Fallback: more informative than bare cluster number
  TRUE ~
    paste(
      "Mixed Conditions ---",
      gsub("_", " ", dominant_weather_group),
      gsub("_", " ", dominant_hour_band)
    )
)]

scenario_final_names[, final_name := ifelse(
  high_risk_archetype,
  paste(final_name, "(High Risk)"),
  final_name
)]

duplicate_scenario_names <- scenario_final_names[, .N, by = final_name][N > 1, final_name]
if (length(duplicate_scenario_names) > 0) {
  scenario_final_names[final_name %in% duplicate_scenario_names, final_name := paste0(final_name, " (S", scenario_cluster, ")")]
}

spatial_final_names <- rbind(
  data.table(
    spatial_cluster = 0L,
    corridor_label = "Noise / Unassigned",
    final_name = "Noise / Unassigned"
  ),
  non_noise_assignments[, .(
    spatial_cluster,
    corridor_label,
    final_name = ifelse(
      corridor_label == "Surface Road Cluster",
      paste("Surface Road Cluster", spatial_cluster),
      paste(corridor_label, "Cluster", spatial_cluster)
    )
  )]
)

scenario_name_map <- scenario_final_names$final_name
names(scenario_name_map) <- scenario_final_names$scenario_cluster

spatial_name_map <- spatial_final_names$final_name
names(spatial_name_map) <- spatial_final_names$spatial_cluster

df_clustered[, scenario_cluster_name := scenario_name_map[as.character(scenario_cluster)]]
df_clustered[, spatial_cluster_name := spatial_name_map[as.character(spatial_cluster)]]

if ("working_name" %in% names(spatial_scenario_crosstab)) {
  spatial_scenario_crosstab[, working_name := NULL]
}
spatial_scenario_crosstab <- merge(
  spatial_scenario_crosstab,
  scenario_final_names[, .(scenario_cluster, final_name)],
  by = "scenario_cluster",
  all.x = TRUE
)

spatial_scenario_dominance <- merge(
  spatial_scenario_dominance[, !c("dominant_scenario_name"), with = FALSE],
  scenario_final_names[, .(scenario_cluster, final_name)],
  by.x = "dominant_scenario_cluster",
  by.y = "scenario_cluster",
  all.x = TRUE
)
data.table::setnames(spatial_scenario_dominance, "final_name", "dominant_scenario_name")

fwrite(spatial_scenario_crosstab, file.path(output_dir, "phase8_spatial_scenario_crosstab.csv"))
fwrite(spatial_scenario_dominance, file.path(output_dir, "phase8_spatial_scenario_dominance.csv"))

fwrite(scenario_final_names, file.path(output_dir, "phase8_scenario_final_names.csv"))
fwrite(spatial_final_names, file.path(output_dir, "phase8_spatial_final_names.csv"))

knitr::kable(scenario_final_names, caption = "Phase 8 final scenario cluster names")
knitr::kable(spatial_final_names, caption = "Phase 8 final spatial cluster names")
```

```{r phase-8-summary}
phase_8_summary <- data.table(
  checkpoint = c(
    "Label join saved",
    "Severity profile saved",
    "County profile saved",
    "Weather-time profile saved",
    "Spatial-scenario cross-tab saved",
    "Final cluster names saved",
    "Scenario names are unique"
  ),
  status = c(
    file.exists(file.path(output_dir, "phase8_label_join_check.csv")),
    file.exists(file.path(output_dir, "phase8_scenario_severity_profile.csv")),
    file.exists(file.path(output_dir, "phase8_scenario_county_profile.csv")),
    file.exists(file.path(output_dir, "phase8_scenario_weather_time_profile.csv")),
    file.exists(file.path(output_dir, "phase8_spatial_scenario_crosstab.csv")),
    file.exists(file.path(output_dir, "phase8_scenario_final_names.csv")) &
      file.exists(file.path(output_dir, "phase8_spatial_final_names.csv")),
    uniqueN(scenario_final_names$final_name) == nrow(scenario_final_names)
  )
)

fwrite(phase_8_summary, file.path(output_dir, "phase8_summary.csv"))

knitr::kable(phase_8_summary, caption = "Phase 8 readiness summary")
```

# Phase 9: Export and Visualization

This phase organizes the report-ready figures, tables, and labeled dataset export so the final write-up can pull from a single output location.

```{r phase-9-export-setup}
figure_dir <- file.path(output_dir, "figures")
table_dir <- file.path(output_dir, "tables")
interactive_dir <- file.path(output_dir, "interactive")

dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(interactive_dir, recursive = TRUE, showWarnings = FALSE)

export_setup <- data.table(
  item = c("Figure directory", "Table directory", "Interactive directory"),
  path = c(figure_dir, table_dir, interactive_dir),
  status = c(
    "Created",
    "Created",
    ifelse(interactive_exports_enabled, "Created", "Created for optional HTML exports")
  )
)

knitr::kable(export_setup, caption = "Phase 9 export folders")
```

```{r phase-9-validation-plot, fig.height=4.5, fig.width=7}
selected_validation_point <- validation_table[k == selected_k]
validation_subtitle <- if (phase6_validation_mode == "consensus_sampled") {
  paste(
    "Consensus silhouette validation across",
    phase6_validation_repeats,
    "repeats of",
    validation_rows,
    "rows"
  )
} else if (phase6_validation_mode == "sampled") {
  paste("Sampled silhouette validation on", validation_rows, "rows")
} else {
  "Full-data silhouette validation"
}

validation_plot <- ggplot2::ggplot(
  validation_table,
  ggplot2::aes(x = k, y = silhouette)
) +
  # v2.0: 0.30 threshold band so reader sees where results sit
  ggplot2::annotate(
    "rect", xmin = -Inf, xmax = Inf,
    ymin = 0.30, ymax = Inf, alpha = 0.06, fill = "#31a354"
  ) +
  ggplot2::annotate(
    "rect", xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = 0.30, alpha = 0.06, fill = "#fb6a4a"
  ) +
  ggplot2::geom_hline(
    yintercept = 0.30, linetype = "dashed",
    color = "#e31a1c", linewidth = 0.8
  ) +
  ggplot2::annotate(
    "text", x = max(validation_table$k),
    y = 0.305, label = "Threshold 0.30",
    hjust = 1, size = 2.9, color = "#e31a1c"
  ) +
  ggplot2::geom_line(color = "#1f78b4", linewidth = 0.9) +
  ggplot2::geom_point(color = "#1f78b4", size = 2.6) +
  ggplot2::geom_vline(
    xintercept = selected_k, linetype = "dashed",
    color = "#33a02c", linewidth = 0.8
  ) +
  ggplot2::geom_point(
    data = selected_validation_point,
    color = "#33a02c", size = 4.5
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = round(silhouette, 3)),
    nudge_y = max(validation_table$silhouette) * 0.03,
    size = 3.2, color = "#1f78b4"
  ) +
  ggplot2::annotate(
    "label",
    x = selected_k,
    y = selected_validation_point$silhouette[[1]],
    label = paste0("Selected k = ", selected_k,
                   "\nSilhouette = ",
                   round(selected_validation_point$silhouette[[1]], 3)),
    fill   = "#e5f5e0", color = "#006d2c",
    label.size = 0.25, alpha = 0.95, size = 3
  ) +
  ggplot2::labs(
    title    = "Silhouette Validation Index by k",
    subtitle = validation_subtitle,
    x        = "Number of clusters (k)",
    y        = "Silhouette width"
  ) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

print(validation_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_validation_index_plot.png"),
  plot = validation_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

```{r phase-9-stability-plot, fig.height=4.5, fig.width=7}
stability_plot_data <- data.table::melt(
  copy(stability_summary),
  id.vars = "k",
  measure.vars = c("rand", "jaccard"),
  variable.name = "metric",
  value.name = "score"
)

# v2.0: Quality-aware subtitle — explicitly flags red-zone result
stability_quality_label <- if (!sil_pass || !jaccard_pass) {
  paste0(
    "WARNING: EXPLORATORY results. ",
    "Jaccard = ", round(selected_jaccard_v2, 3),
    " (threshold 0.40). ",
    "Silhouette = ", round(selected_silhouette, 3),
    " (threshold 0.30)."
  )
} else {
  paste0(
    "Cluster quality meets thresholds. ",
    "Jaccard = ", round(selected_jaccard_v2, 3),
    ", Silhouette = ", round(selected_silhouette, 3), "."
  )
}

stability_plot <- ggplot2::ggplot(
  stability_plot_data,
  ggplot2::aes(x = factor(k), y = score, fill = metric)
) +
  ggplot2::annotate(
    "rect", xmin = -Inf, xmax = Inf,
    ymin = 0.75, ymax = 1.05, alpha = 0.08, fill = "#31a354"
  ) +
  ggplot2::annotate(
    "rect", xmin = -Inf, xmax = Inf,
    ymin = 0.60, ymax = 0.75, alpha = 0.08, fill = "#fec44f"
  ) +
  ggplot2::annotate(
    "rect", xmin = -Inf, xmax = Inf,
    ymin = 0.00, ymax = 0.60, alpha = 0.06, fill = "#fb6a4a"
  ) +
  ggplot2::geom_col(position = "dodge", width = 0.65) +
  ggplot2::geom_text(
    ggplot2::aes(label = round(score, 3)),
    position = ggplot2::position_dodge(width = 0.65),
    vjust = -0.4, size = 3.2, fontface = "bold"
  ) +
  ggplot2::geom_hline(
    yintercept = 0.40, linetype = "dashed",
    color = "#e31a1c", linewidth = 0.8
  ) +
  ggplot2::geom_hline(
    yintercept = 0.60, linetype = "dashed",
    color = "#fd8d3c", linewidth = 0.7
  ) +
  ggplot2::geom_hline(
    yintercept = 0.75, linetype = "dotted",
    color = "#238b45", linewidth = 0.7
  ) +
  ggplot2::annotate(
    "text", x = Inf, y = 0.41,
    label = "Jaccard threshold 0.40",
    hjust = 1.05, size = 2.8, color = "#e31a1c"
  ) +
  ggplot2::annotate(
    "text", x = Inf, y = 0.76,
    label = "Stable threshold 0.75",
    hjust = 1.05, size = 2.8, color = "#238b45"
  ) +
  ggplot2::scale_fill_manual(
    values = c("rand" = "#4292c6", "jaccard" = "#e6550d"),
    labels = c("rand" = "Rand index", "jaccard" = "Jaccard index")
  ) +
  ggplot2::labs(
    title    = "Cluster Stability — Rand and Jaccard by k",
    subtitle = stability_quality_label,
    x        = "Number of clusters (k)",
    y        = "Stability score",
    fill     = "Metric"
  ) +
  ggplot2::coord_cartesian(ylim = c(0, 1.05)) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    plot.title    = ggplot2::element_text(face = "bold"),
    plot.subtitle = ggplot2::element_text(
      color = if (!sil_pass || !jaccard_pass) "#cb181d" else "#238b45",
      size  = 9
    ),
    panel.grid.minor = ggplot2::element_blank()
  )

print(stability_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_stability_plot.png"),
  plot = stability_plot,
  width = 7,
  height = 4.5,
  dpi = 300
)
```

```{r phase-9-spatial-scenario-heatmap, fig.height=5.5, fig.width=8}
heatmap_data <- merge(
  spatial_scenario_crosstab,
  scenario_final_names[, .(scenario_cluster, scenario_name_short = wrap_label(final_name, width = 18))],
  by = "scenario_cluster",
  all.x = TRUE
)

heatmap_data <- merge(
  heatmap_data,
  spatial_final_names[, .(spatial_cluster, spatial_name_short = wrap_label(final_name, width = 20))],
  by = "spatial_cluster",
  all.x = TRUE
)

heatmap_plot <- ggplot2::ggplot(
  heatmap_data,
  ggplot2::aes(
    x = factor(scenario_cluster),
    y = factor(spatial_cluster),
    fill = spatial_cluster_pct
  )
) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::geom_text(ggplot2::aes(label = round(spatial_cluster_pct, 1)), size = 3) +
  ggplot2::scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  ggplot2::labs(
    title = "Phase 9 Spatial-Scenario Cross-Tab Heatmap",
    x = "Scenario cluster",
    y = "Spatial cluster",
    fill = "% within\nspatial cluster"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    panel.grid = ggplot2::element_blank()
  )

print(heatmap_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_spatial_scenario_heatmap.png"),
  plot = heatmap_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

```{r phase-9-spatial-cluster-facets, fig.height=7, fig.width=9}
top_spatial_clusters <- non_noise_assignments[order(-cluster_size)][1:min(.N, 6), spatial_cluster]

spatial_cluster_heatmap_data <- merge(
  df_spatial_plot[spatial_cluster %in% top_spatial_clusters],
  spatial_final_names[, .(spatial_cluster, final_name)],
  by = "spatial_cluster",
  all.x = TRUE
)

spatial_cluster_heatmap_data[, final_name_short := wrap_label(final_name, width = 26)]

spatial_cluster_facet_heatmap <- ggplot2::ggplot(
  spatial_cluster_heatmap_data,
  ggplot2::aes(
    x = x_km,
    y = y_km,
    fill = after_stat(log10(count + 1))
  )
) +
  ggplot2::geom_bin2d(bins = 50) +
  ggplot2::scale_fill_viridis_c(
    option = "D",
    name = "log10(bin\ncount + 1)"
  ) +
  ggplot2::coord_equal() +
  ggplot2::facet_wrap(~ final_name_short, ncol = 2) +
  ggplot2::labs(
    title = "Phase 9 Spatial Cluster Heatmaps",
    subtitle = "Each panel shows the internal density shape of one of the largest hotspot clusters",
    x = "Projected x (km)",
    y = "Projected y (km)"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )

print(spatial_cluster_facet_heatmap)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_spatial_cluster_facets_heatmap.png"),
  plot = spatial_cluster_facet_heatmap,
  width = 9,
  height = 7,
  dpi = 300
)
```

```{r phase-9-scenario-risk-bubble, fig.height=5, fig.width=7.5}
dataset_sev34_pct <- round(100 * mean(as.integer(df_clustered$Severity) %in% c(3, 4)), 2)

risk_bubble_data <- merge(
  scenario_severity_profile,
  scenario_weather_time_profile[, .(
    scenario_cluster,
    dominant_weather_group,
    dominant_hour_band,
    mean_visibility
  )],
  by = "scenario_cluster",
  all.x = TRUE
)

risk_bubble_data <- merge(
  risk_bubble_data,
  scenario_final_names[, .(scenario_cluster, final_name)],
  by = "scenario_cluster",
  all.x = TRUE
)

# v2.0: Use truncated final name as label; add sev4_vs_baseline annotation
risk_bubble_data[, short_label := paste0(
  "S", scenario_cluster, ": ",
  substr(gsub(" ---.*", "", final_name), 1, 18)
)]
risk_bubble_data <- merge(
  risk_bubble_data,
  scenario_severity_profile[, .(scenario_cluster, sev4_vs_baseline)],
  by = "scenario_cluster", all.x = TRUE
)

risk_bubble_plot <- ggplot2::ggplot(
  risk_bubble_data,
  ggplot2::aes(
    x     = n,
    y     = sev34_pct,
    size  = pmax(sev4_pct, 0.1),
    color = dominant_weather_group
  )
) +
  ggplot2::geom_hline(
    yintercept = dataset_sev34_pct, linetype = "dashed",
    color = "#e31a1c", linewidth = 0.8
  ) +
  ggplot2::annotate(
    "text", x = -Inf, y = dataset_sev34_pct + 0.3,
    label = paste("Dataset baseline:", dataset_sev34_pct, "%"),
    hjust = -0.05, size = 2.9, color = "#e31a1c"
  ) +
  ggplot2::geom_point(alpha = 0.82) +
  ggrepel::geom_text_repel(
    ggplot2::aes(label = short_label),
    size = 3, show.legend = FALSE,
    box.padding = 0.4, max.overlaps = 10
  ) +
  ggplot2::scale_size_continuous(range = c(4, 14)) +
  ggplot2::scale_color_brewer(palette = "Set1") +
  ggplot2::labs(
    title    = "Scenario Cluster Risk and Size Profile",
    subtitle = paste0(
      "Bubble size = Sev4 share. Dashed line = dataset Sev3/4 baseline (",
      dataset_sev34_pct, "%). Label shows cluster name prefix."
    ),
    x     = "Cluster size (n)",
    y     = "Severity 3/4 share (%)",
    size  = "Sev4 share (%)",
    color = "Weather group"
  ) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    legend.position  = "right"
  )

print(risk_bubble_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_scenario_risk_bubble.png"),
  plot = risk_bubble_plot,
  width = 7.5,
  height = 5,
  dpi = 300
)
```

```{r phase-9-county-composition, fig.height=4.8, fig.width=7.5}
county_plot_data <- data.table::melt(
  copy(scenario_county_profile),
  id.vars = c("scenario_cluster", "n", "dominant_county"),
  measure.vars = c("Nassau_pct", "Suffolk_pct", "Queens_pct"),
  variable.name = "county",
  value.name = "pct"
)

county_plot_data[, county := factor(
  county,
  levels = c("Nassau_pct", "Suffolk_pct", "Queens_pct"),
  labels = c("Nassau", "Suffolk", "Queens")
)]

county_plot <- ggplot2::ggplot(
  county_plot_data,
  ggplot2::aes(x = factor(scenario_cluster), y = pct, fill = county)
) +
  ggplot2::geom_col(width = 0.75) +
  ggplot2::labs(
    title = "Phase 9 County Composition by Scenario Cluster",
    subtitle = "Each bar sums to 100% within a scenario cluster",
    x = "Scenario cluster",
    y = "Share (%)",
    fill = "County"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

print(county_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_county_composition.png"),
  plot = county_plot,
  width = 7.5,
  height = 4.8,
  dpi = 300
)
```

```{r phase-9-cluster-fingerprint, fig.height=5.8, fig.width=8}
# v2.0: Add sev4_rate row; use final cluster names on x-axis
fingerprint_source <- merge(
  scenario_weather_time_profile[, .(
    scenario_cluster, night_pct, weekend_pct,
    junction_pct, traffic_signal_pct, crossing_pct,
    mean_visibility, mean_temperature, mean_wind_speed
  )],
  scenario_severity_profile[, .(scenario_cluster, sev34_pct, sev4_rate)],
  by = "scenario_cluster", all.x = TRUE
)

# Attach short final names for x-axis labels
fingerprint_source <- merge(
  fingerprint_source,
  scenario_final_names[, .(
    scenario_cluster,
    cluster_label = paste0(
      "S", scenario_cluster, "\n",
      substr(gsub(" ---.*", "", final_name), 1, 16)
    )
  )],
  by = "scenario_cluster", all.x = TRUE
)

fingerprint_long <- data.table::melt(
  fingerprint_source,
  id.vars       = c("scenario_cluster", "cluster_label"),
  variable.name = "metric",
  value.name    = "metric_value"
)

metric_labels <- c(
  night_pct          = "Night share (%)",
  weekend_pct        = "Weekend share (%)",
  junction_pct       = "Junction share (%)",
  traffic_signal_pct = "Signal share (%)",
  crossing_pct       = "Crossing share (%)",
  mean_visibility    = "Mean visibility (mi)",
  mean_temperature   = "Mean temperature (F)",
  mean_wind_speed    = "Mean wind speed (mph)",
  sev34_pct          = "Severity 3/4 share (%)",
  sev4_rate          = "Severity 4 rate (%)"
)

fingerprint_long[, metric_label := factor(
  metric_labels[as.character(metric)],
  levels = rev(metric_labels)
)]
fingerprint_long[, scaled_value := rescale01(metric_value), by = metric]

fingerprint_plot <- ggplot2::ggplot(
  fingerprint_long,
  ggplot2::aes(
    x    = reorder(cluster_label, scenario_cluster),
    y    = metric_label,
    fill = scaled_value
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(label = round(metric_value, 1)),
    size = 3, fontface = "bold"
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#f7fbff", mid = "#6baed6", high = "#08306b",
    midpoint = 0.5, na.value = "grey90"
  ) +
  ggplot2::labs(
    title    = "Scenario Cluster Fingerprint Heatmap",
    subtitle = "Tile color scaled within each metric row; labels show actual values",
    x        = "Scenario cluster",
    y        = NULL,
    fill     = "Relative\nstrength"
  ) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid       = ggplot2::element_blank(),
    axis.text.x      = ggplot2::element_text(size = 8, lineheight = 1.1),
    axis.text.y      = ggplot2::element_text(size = 9)
  )

print(fingerprint_plot)

ggplot2::ggsave(
  filename = file.path(output_dir, "phase9_cluster_fingerprint_heatmap.png"),
  plot = fingerprint_plot,
  width = 8,
  height = 5.8,
  dpi = 300
)
```

```{r phase-9-export-artifacts}
write.csv(
  df_clustered,
  file.path(output_dir, "li_accidents_clustered.csv"),
  row.names = FALSE
)

table_exports <- list(
  phase6_validation_indices = validation_table,
  phase6_validation_metadata = validation_metadata,
  phase6_stability_summary = stability_summary,
  phase6_selection_summary = selection_summary,
  phase7_eps_trials = eps_trials,
  phase7_parameter_summary = parameter_summary,
  phase7_dbscan_cluster_sizes = spatial_cluster_sizes,
  phase7_corridor_assignments = non_noise_assignments,
  phase8_severity_profile = scenario_severity_profile,
  phase8_county_profile = scenario_county_profile,
  phase8_weather_time_profile = scenario_weather_time_profile,
  phase8_spatial_scenario_crosstab = spatial_scenario_crosstab,
  phase8_spatial_scenario_dominance = spatial_scenario_dominance,
  phase8_scenario_final_names = scenario_final_names,
  phase8_spatial_final_names = spatial_final_names
)

if (!is.null(validation_repeat_results)) {
  table_exports$phase6_validation_repeats <- validation_repeat_results
}

if (!is.null(stability_repeat_results)) {
  table_exports$phase6_stability_repeats <- stability_repeat_results
}

invisible(lapply(names(table_exports), function(table_name) {
  fwrite(table_exports[[table_name]], file.path(table_dir, paste0(table_name, ".csv")))
}))

figure_exports <- c(
  "phase5_kproto_elbow_plot.png",
  "phase6_cluster_profiles.pdf",
  "phase7_knn_distance_plot.png",
  "phase7_spatial_hotspot_map.png",
  "phase7_spatial_density_heatmap.png",
  "phase9_validation_index_plot.png",
  "phase9_stability_plot.png",
  "phase9_spatial_scenario_heatmap.png",
  "phase9_spatial_cluster_facets_heatmap.png",
  "phase9_scenario_risk_bubble.png",
  "phase9_county_composition.png",
  "phase9_cluster_fingerprint_heatmap.png"
)

invisible(lapply(figure_exports, function(file_name) {
  source_path <- file.path(output_dir, file_name)
  if (file.exists(source_path)) {
    file.copy(source_path, file.path(figure_dir, file_name), overwrite = TRUE)
  }
}))

file.copy(
  file.path(output_dir, "li_accidents_clustered.csv"),
  file.path(table_dir, "li_accidents_clustered.csv"),
  overwrite = TRUE
)

export_manifest <- data.table(
  artifact_type = c(
    rep("figure", length(figure_exports)),
    rep("table", length(table_exports)),
    "table"
  ),
  artifact_name = c(
    figure_exports,
    paste0(names(table_exports), ".csv"),
    "li_accidents_clustered.csv"
  ),
  target_path = c(
    file.path(figure_dir, figure_exports),
    file.path(table_dir, paste0(names(table_exports), ".csv")),
    file.path(table_dir, "li_accidents_clustered.csv")
  )
)

fwrite(export_manifest, file.path(output_dir, "phase9_export_manifest.csv"))

knitr::kable(export_manifest, caption = "Phase 9 export manifest")
```

```{r phase-9-feature-sensitivity-plot, fig.height=4.5, fig.width=7.5}
# v2.0: Side-by-side silhouette comparison — full vs reduced feature set.
# Shows whether dropping imputed features (Pressure, Visibility) improves quality.
if (file.exists(
  file.path(output_dir, "phase6_feature_silhouette_comparison.csv")
)) {
  feat_compare <- data.table::fread(
    file.path(output_dir, "phase6_feature_silhouette_comparison.csv")
  )

  feat_compare[, feature_set_label := ifelse(
    feature_set == "full",
    "Full (16 features)",
    "Reduced (14 features,\nno Pressure/Visibility)"
  )]

  feature_sensitivity_plot <- ggplot2::ggplot(
    feat_compare,
    ggplot2::aes(
      x    = factor(k),
      y    = silhouette,
      fill = feature_set_label
    )
  ) +
    ggplot2::geom_col(position = "dodge", width = 0.65) +
    ggplot2::geom_text(
      ggplot2::aes(label = round(silhouette, 3)),
      position = ggplot2::position_dodge(width = 0.65),
      vjust = -0.4, size = 3, fontface = "bold"
    ) +
    ggplot2::geom_hline(
      yintercept = 0.30, linetype = "dashed",
      color = "#e31a1c", linewidth = 0.8
    ) +
    ggplot2::annotate(
      "text", x = Inf, y = 0.305,
      label = "Threshold 0.30",
      hjust = 1.05, size = 2.8, color = "#e31a1c"
    ) +
    ggplot2::scale_fill_manual(
      values = c(
        "Full (16 features)"                        = "#4292c6",
        "Reduced (14 features,\nno Pressure/Visibility)" = "#e6550d"
      )
    ) +
    ggplot2::labs(
      title    = "Feature Sensitivity: Silhouette by k",
      subtitle = "Reduced set drops Pressure(in) and Visibility(mi) — both were median-imputed",
      x        = "Number of clusters (k)",
      y        = "Silhouette width",
      fill     = "Feature set"
    ) +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold"),
      panel.grid.minor = ggplot2::element_blank(),
      legend.position  = "bottom"
    )

  print(feature_sensitivity_plot)

  ggplot2::ggsave(
    filename = file.path(
      output_dir, "figures", "phase9_feature_sensitivity.png"
    ),
    plot   = feature_sensitivity_plot,
    width  = 7.5, height = 4.5, dpi = 300
  )
} else {
  cat("Feature silhouette comparison not available — skipping plot.\n")
}
```

```{r phase-9-interactive-exports}
interactive_exports <- data.table(
  artifact_name = c(
    "phase9_validation_index_interactive.html",
    "phase9_stability_interactive.html",
    "phase9_spatial_hotspot_interactive.html",
    "phase9_spatial_density_interactive.html",
    "phase9_scenario_risk_interactive.html",
    "phase9_county_composition_interactive.html",
    "phase9_cluster_fingerprint_interactive.html",
    "phase9_spatial_scenario_heatmap_interactive.html"
  ),
  target_path = file.path(
    interactive_dir,
    c(
      "phase9_validation_index_interactive.html",
      "phase9_stability_interactive.html",
      "phase9_spatial_hotspot_interactive.html",
      "phase9_spatial_density_interactive.html",
      "phase9_scenario_risk_interactive.html",
      "phase9_county_composition_interactive.html",
      "phase9_cluster_fingerprint_interactive.html",
      "phase9_spatial_scenario_heatmap_interactive.html"
    )
  ),
  saved = FALSE,
  note = ifelse(
    interactive_exports_enabled,
    "Saved as interactive HTML",
    "Install plotly and htmlwidgets to enable interactive exports"
  )
)

if (interactive_exports_enabled) {
  spatial_interactive_data <- copy(df_spatial_plot)
  if (nrow(spatial_interactive_data) > 15000) {
    set.seed(42)
    spatial_interactive_data <- spatial_interactive_data[
      ,
      .SD[sample(.N, size = min(.N, max(1L, ceiling(15000 * .N / nrow(df_spatial_plot)))))],
      by = cluster_label
    ]
  }

  spatial_map_interactive <- ggplot2::ggplot(
    spatial_interactive_data,
    ggplot2::aes(x = x_km, y = y_km, color = cluster_label,
                 text = paste0(
                   "Cluster: ", cluster_label,
                   "<br>x_km: ", round(x_km, 2),
                   "<br>y_km: ", round(y_km, 2)
                 ))
  ) +
    ggplot2::geom_point(size = 0.6, alpha = 0.7) +
    ggplot2::scale_color_manual(values = cluster_colors) +
    ggplot2::labs(
      title = "Interactive Spatial Hotspot View",
      subtitle = "Rendered on a representative sample for smoother browsing",
      x = "Projected x (km)",
      y = "Projected y (km)",
      color = "Cluster"
    ) +
    ggplot2::theme_minimal()

  interactive_widgets <- list(
    phase9_validation_index_interactive.html = plotly::ggplotly(validation_plot, tooltip = c("x", "y")),
    phase9_stability_interactive.html = plotly::ggplotly(stability_plot, tooltip = c("x", "y", "fill")),
    phase9_spatial_hotspot_interactive.html = plotly::ggplotly(spatial_map_interactive, tooltip = "text"),
    phase9_spatial_density_interactive.html = plotly::plot_ly(
      spatial_interactive_data,
      x = ~x_km,
      y = ~y_km,
      type = "histogram2d",
      colorscale = "Viridis",
      nbinsx = 80,
      nbinsy = 60
    ) |>
      plotly::layout(
        title = "Interactive Spatial Density Heatmap",
        xaxis = list(title = "Projected x (km)"),
        yaxis = list(title = "Projected y (km)")
      ),
    phase9_scenario_risk_interactive.html = plotly::ggplotly(risk_bubble_plot, tooltip = c("x", "y", "size", "colour")),
    phase9_county_composition_interactive.html = plotly::ggplotly(county_plot, tooltip = c("x", "y", "fill")),
    phase9_cluster_fingerprint_interactive.html = plotly::ggplotly(fingerprint_plot, tooltip = c("x", "y", "fill")),
    phase9_spatial_scenario_heatmap_interactive.html = plotly::ggplotly(heatmap_plot, tooltip = c("x", "y", "fill"))
  )

  invisible(lapply(names(interactive_widgets), function(file_name) {
    htmlwidgets::saveWidget(
      interactive_widgets[[file_name]],
      file = file.path(interactive_dir, file_name),
      selfcontained = FALSE
    )
  }))

  interactive_exports[, saved := file.exists(target_path)]
}

fwrite(interactive_exports, file.path(output_dir, "phase9_interactive_manifest.csv"))

knitr::kable(interactive_exports, caption = "Phase 9 interactive export manifest")
```

```{r phase-9-summary}
phase_9_summary <- data.table(
  checkpoint = c(
    "Figure directory created",
    "Table directory created",
    "Validation plot saved",
    "Stability plot saved",
    "Heatmap saved",
    "Spatial density heatmap saved",
    "Spatial cluster facet heatmap saved",
    "Scenario risk bubble saved",
    "County composition plot saved",
    "Cluster fingerprint heatmap saved",
    "Labeled CSV exported",
    "Interactive manifest saved",
    "Export manifest saved"
  ),
  status = c(
    dir.exists(figure_dir),
    dir.exists(table_dir),
    file.exists(file.path(figure_dir, "phase9_validation_index_plot.png")),
    file.exists(file.path(figure_dir, "phase9_stability_plot.png")),
    file.exists(file.path(figure_dir, "phase9_spatial_scenario_heatmap.png")),
    file.exists(file.path(figure_dir, "phase7_spatial_density_heatmap.png")),
    file.exists(file.path(figure_dir, "phase9_spatial_cluster_facets_heatmap.png")),
    file.exists(file.path(figure_dir, "phase9_scenario_risk_bubble.png")),
    file.exists(file.path(figure_dir, "phase9_county_composition.png")),
    file.exists(file.path(figure_dir, "phase9_cluster_fingerprint_heatmap.png")),
    file.exists(file.path(table_dir, "li_accidents_clustered.csv")),
    file.exists(file.path(output_dir, "phase9_interactive_manifest.csv")),
    file.exists(file.path(output_dir, "phase9_export_manifest.csv"))
  )
)

fwrite(phase_9_summary, file.path(output_dir, "phase9_summary.csv"))

knitr::kable(phase_9_summary, caption = "Phase 9 readiness summary")
```

# Phase 10: Final Report and Submission Checklist

This phase keeps a simple readiness tracker for the final report assets and submission handoff.

```{r phase-10-readiness}
phase_10_checklist <- data.table(
  deliverable = c(
    "Long Island SRS source",
    "Execution plan source",
    "Main workflow Rmd",
    "Cleaned dataset",
    "Clustered labeled dataset",
    "Figure export folder",
    "Table export folder",
    "Interactive export manifest"
  ),
  path = c(
    "Li_Clustering_SRS.tex",
    "Li_Clustering_Execution_Plan.tex",
    "li_clustering.Rmd",
    input_path,
    file.path(table_dir, "li_accidents_clustered.csv"),
    figure_dir,
    table_dir,
    file.path(output_dir, "phase9_interactive_manifest.csv")
  ),
  ready = c(
    file.exists("Li_Clustering_SRS.tex"),
    file.exists("Li_Clustering_Execution_Plan.tex"),
    file.exists("li_clustering.Rmd"),
    file.exists(input_path),
    file.exists(file.path(table_dir, "li_accidents_clustered.csv")),
    dir.exists(figure_dir),
    dir.exists(table_dir),
    file.exists(file.path(output_dir, "phase9_interactive_manifest.csv"))
  )
)

fwrite(phase_10_checklist, file.path(output_dir, "phase10_readiness_checklist.csv"))

knitr::kable(phase_10_checklist, caption = "Phase 10 readiness checklist")
```

```{r phase-10-report-outline}
phase_10_report_outline <- data.table(
  report_section = c(
    "Introduction and motivation",
    "Dataset description and audit findings",
    "Feature engineering decisions",
    "Method selection rationale",
    "Scenario clustering results",
    "Spatial hotspot results",
    "Interpretation and cross-tab findings",
    "Limitations",
    "Conclusion and safety priorities"
  ),
  main_source = c(
    "Li_Clustering_SRS.tex",
    file.path(table_dir, "phase8_severity_profile.csv"),
    file.path(table_dir, "phase8_weather_time_profile.csv"),
    "Li_Clustering_Execution_Plan.tex",
    file.path(figure_dir, "phase9_validation_index_plot.png"),
    file.path(figure_dir, "phase7_spatial_hotspot_map.png"),
    file.path(table_dir, "phase8_spatial_scenario_crosstab.csv"),
    file.path(output_dir, "phase10_limitations_table.csv"),
    file.path(output_dir, "phase10_research_question_coverage.csv")
  ),
  ready = c(
    file.exists("Li_Clustering_SRS.tex"),
    file.exists(file.path(table_dir, "phase8_severity_profile.csv")),
    file.exists(file.path(table_dir, "phase8_weather_time_profile.csv")),
    file.exists("Li_Clustering_Execution_Plan.tex"),
    file.exists(file.path(figure_dir, "phase9_validation_index_plot.png")),
    file.exists(file.path(figure_dir, "phase7_spatial_hotspot_map.png")),
    file.exists(file.path(table_dir, "phase8_spatial_scenario_crosstab.csv")),
    TRUE,
    TRUE
  )
)

fwrite(phase_10_report_outline, file.path(output_dir, "phase10_report_outline.csv"))

knitr::kable(phase_10_report_outline, caption = "Phase 10 report outline and source map")
```

```{r phase-10-limitations}
# v2.0: Cluster quality failure rows prepended FIRST so they appear at the
# top of the rendered table and cannot be missed in the final report.
cluster_quality_rows <- data.table(
  limitation = c(
    "Scenario cluster quality below standard thresholds",
    "Scenario clustering is exploratory only"
  ),
  report_language = c(
    paste0(
      "Silhouette width = ", round(selected_silhouette, 3),
      " (published threshold >= 0.30; Kaufman and Rousseeuw 1990). ",
      "Jaccard stability index = ", round(selected_jaccard_v2, 3),
      " (threshold >= 0.40 for exploratory, >= 0.60 for acceptable; ",
      "Hennig 2007). Both metrics fall below acceptable thresholds, ",
      "indicating that cluster boundaries are weak and that bootstrap ",
      "resamples do not reliably reproduce the same partition."
    ),
    paste0(
      "Because silhouette and Jaccard are both below threshold, ",
      "the k = ", selected_k, " scenario clustering solution should be ",
      "interpreted as a set of candidate patterns for hypothesis generation, ",
      "not as stable or confirmed accident archetypes. ",
      "The elevation of Severity 3/4 in Clusters 4 and 5 is a plausible ",
      "signal but requires replication on a new sample before it can be ",
      "treated as a reliable finding. ",
      "DBSCAN spatial results are separately validated and remain defensible."
    )
  ),
  include_in_final_report = TRUE
)

phase_10_limitations <- rbind(
  cluster_quality_rows,
  data.table(
    limitation = c(
      "Duration capping",
      "Precipitation exclusion",
      "Distance exclusion",
      "Severity 2 dominance",
      "Sampled Phase 6 validation",
      "DBSCAN parameter sensitivity"
    ),
    report_language = c(
      "Duration was capped at the 99th percentile to reduce the effect of extreme outliers, so the top 1% of incidents lose some real variation.",
      "Precipitation was excluded from the core model because missingness was high enough (20.1%) to make routine imputation less defensible.",
      "Distance(mi) was excluded because the large share of zero values suggests a reporting artifact rather than a stable behavioral feature.",
      "Severity 2 dominates the dataset (85% of rows), so subtle differences among severe crashes can be harder to separate.",
      if (run_profile == "full_srs") {
        "Phase 6 is configured for full-data validation on a stronger machine."
      } else if (run_profile == "consensus_16gb") {
        paste(
          "Phase 6 uses repeated sampled validation (4 x 4,000 rows) and",
          "repeated stability checks (3 x 2,500 rows, 10 bootstraps each)",
          "to build a consensus k on a 16 GB machine."
        )
      } else {
        "Phase 6 uses single-sample silhouette (5,000 rows) and stability (3,000 rows) checks; full-data version archived for high-memory rerun."
      },
      "Spatial hotspot shapes can shift if eps or minPts changes; corridor labels should be read as data-driven approximations confirmed by centroid inspection."
    ),
    include_in_final_report = TRUE
  )
)

fwrite(phase_10_limitations, file.path(output_dir, "phase10_limitations_table.csv"))

knitr::kable(
  phase_10_limitations,
  caption = paste(
    "Phase 10 limitations table (v2.0).",
    "First two rows document cluster quality failure explicitly."
  )
)
```

```{r phase-10-research-question-coverage}
phase_10_rq_coverage <- data.table(
  research_focus = c(
    "Scenario accident archetypes",
    "Spatial hotspot corridors",
    "High-risk severity patterns",
    "County composition differences",
    "Hotspot-to-scenario linkage"
  ),
  evidence_source = c(
    file.path(table_dir, "phase8_scenario_final_names.csv"),
    file.path(table_dir, "phase8_spatial_final_names.csv"),
    file.path(table_dir, "phase8_severity_profile.csv"),
    file.path(table_dir, "phase8_county_profile.csv"),
    file.path(table_dir, "phase8_spatial_scenario_crosstab.csv")
  ),
  ready = c(
    file.exists(file.path(table_dir, "phase8_scenario_final_names.csv")),
    file.exists(file.path(table_dir, "phase8_spatial_final_names.csv")),
    file.exists(file.path(table_dir, "phase8_severity_profile.csv")),
    file.exists(file.path(table_dir, "phase8_county_profile.csv")),
    file.exists(file.path(table_dir, "phase8_spatial_scenario_crosstab.csv"))
  )
)

fwrite(phase_10_rq_coverage, file.path(output_dir, "phase10_research_question_coverage.csv"))

knitr::kable(phase_10_rq_coverage, caption = "Phase 10 research-question coverage")
```

```{r phase-10-submission-manifest}
phase_10_submission_manifest <- data.table(
  deliverable_group = c(
    "Core source",
    "Core source",
    "Core source",
    "Data",
    "Data",
    "Exports",
    "Exports",
    "Exports"
  ),
  deliverable = c(
    "Li_Clustering_SRS.tex",
    "Li_Clustering_Execution_Plan.tex",
    "li_clustering.Rmd",
    "li_accidents_clean.csv",
    "li_accidents_clustered.csv",
    "Figure directory",
    "Table directory",
    "Phase 10 checklist bundle"
  ),
  path = c(
    "Li_Clustering_SRS.tex",
    "Li_Clustering_Execution_Plan.tex",
    "li_clustering.Rmd",
    input_path,
    file.path(table_dir, "li_accidents_clustered.csv"),
    figure_dir,
    table_dir,
    output_dir
  ),
  ready = c(
    file.exists("Li_Clustering_SRS.tex"),
    file.exists("Li_Clustering_Execution_Plan.tex"),
    file.exists("li_clustering.Rmd"),
    file.exists(input_path),
    file.exists(file.path(table_dir, "li_accidents_clustered.csv")),
    dir.exists(figure_dir),
    dir.exists(table_dir),
    dir.exists(output_dir)
  )
)

fwrite(phase_10_submission_manifest, file.path(output_dir, "phase10_submission_manifest.csv"))

knitr::kable(phase_10_submission_manifest, caption = "Phase 10 submission manifest")
```

```{r phase-10-summary}
phase_10_summary <- data.table(
  checkpoint = c(
    "Readiness checklist saved",
    "Report outline saved",
    "Limitations table saved",
    "Research-question coverage saved",
    "Submission manifest saved"
  ),
  status = c(
    file.exists(file.path(output_dir, "phase10_readiness_checklist.csv")),
    file.exists(file.path(output_dir, "phase10_report_outline.csv")),
    file.exists(file.path(output_dir, "phase10_limitations_table.csv")),
    file.exists(file.path(output_dir, "phase10_research_question_coverage.csv")),
    file.exists(file.path(output_dir, "phase10_submission_manifest.csv"))
  )
)

fwrite(phase_10_summary, file.path(output_dir, "phase10_summary.csv"))

knitr::kable(phase_10_summary, caption = "Phase 10 readiness summary")
```

```{r phase-10-sanity-audit}
selected_jaccard_value <- suppressWarnings(as.numeric(selected_jaccard))

phase_10_sanity_audit <- data.table(
  check = c(
    "Scenario names are unique",
    "Scenario labels have no gaps",
    "Spatial labels have no gaps",
    "Selected scenario stability is at least moderate",
    "Largest spatial cluster stays below 60% of all rows",
    "DBSCAN cluster-count target met",
    "DBSCAN noise target met"
  ),
  status = c(
    uniqueN(scenario_final_names$final_name) == nrow(scenario_final_names),
    label_join_check[metric == "Scenario label NAs", value][[1]] == 0,
    label_join_check[metric == "Spatial label NAs", value][[1]] == 0,
    !is.na(selected_jaccard_value) && selected_jaccard_value >= 0.4,
    largest_non_noise_share_pct <= phase7_max_cluster_share_cap,
    uniqueN(df_spatial_clusters$spatial_cluster[df_spatial_clusters$spatial_cluster > 0]) >= 3 &
      uniqueN(df_spatial_clusters$spatial_cluster[df_spatial_clusters$spatial_cluster > 0]) <= 15,
    round(100 * mean(df_spatial_clusters$spatial_cluster == 0), 2) >= 5 &
      round(100 * mean(df_spatial_clusters$spatial_cluster == 0), 2) <= 30
  ),
  note = c(
    "Final scenario names should be distinct for presentation and discussion.",
    "Every row should carry a scenario label after the join.",
    "Every row should carry a spatial label after the join.",
    "A Jaccard score below 0.4 should be presented as exploratory rather than settled.",
    "This avoids one hotspot absorbing most of Long Island.",
    "The SRS target is between 3 and 15 non-noise clusters.",
    "The SRS target is between 5% and 30% noise."
  )
)

fwrite(phase_10_sanity_audit, file.path(output_dir, "phase10_sanity_audit.csv"))

phase_10_summary <- rbind(
  phase_10_summary,
  data.table(
    checkpoint = "Sanity audit saved",
    status = file.exists(file.path(output_dir, "phase10_sanity_audit.csv"))
  )
)
fwrite(phase_10_summary, file.path(output_dir, "phase10_summary.csv"))

knitr::kable(phase_10_sanity_audit, caption = "Phase 10 sanity audit")
```

# Notes for the next phases

- Phase 2 should stay read-only and focus on profiling the cleaned dataset.
- Phase 3 now derives the main engineered features needed for the scenario and spatial matrices.
- Phase 4 now defines the actual scenario and spatial matrices for clustering.
- Phase 5 now sets up the candidate k-prototypes runs and their core diagnostics.
- Phase 6 now validates the candidate solutions, checks stability, and assigns a selected k.
- Phase 7 now sets DBSCAN parameters, creates the spatial hotspot solution, and assigns corridor-style labels.
- Phase 8 now joins cluster labels back to the working data and builds the interpretation tables.
- Phase 9 now exports the report-ready figures, tables, and labeled dataset.
- Phase 10 now tracks final report and submission readiness.
- `run_profile = "consensus_16gb"` is the current default in this file.
- `run_profile = "local_safe"` is the quickest laptop-friendly option.
- `run_profile = "consensus_16gb"` repeats sampled Phase 6 checks and aggregates them into a consensus choice for 16 GB machines.
- `run_profile = "full_srs"` is reserved for a stronger machine that can support the full Phase 6 diagnostics.
- The current default uses repeated sampled validation and repeated sampled stability checks to stay within 16 GB memory limits.
- Interactive HTML exports are optional and now turn on automatically when the needed packages are available.
- The original full-data Phase 6 validation and stability calls are archived in `eval=FALSE` chunks for later reruns.
- Phase 3 and later should write tables, figures, and exports into `local_artifacts/output/`.
- No `.RData` workspace files are needed for this workflow.

```{r session-info}
sessionInfo()
```
